---
title: "Grades"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 4
    fig_caption: yes
    # code_folding: show
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to check your grades
* If the following code returns empty data.tables, then
please double check that you have entered your student ID
correctly **as an integer**.

* If you still get empty data.tables then it may be because
there was a fatal error in your homework submission that
crashed my grading program. Please email me directly to
inquire.

```{r, eval=F}
library(data.table)
library(ggplot2)

rm(list=ls())

# Enter your student ID below as an integer
sid <- 314159

# HW 1
d1 <- fread('https://crossley.github.io/cogs2020/grades/hw_1_table.csv')
d1[student_id == sid]

# HW 2
d2 <- fread('https://crossley.github.io/cogs2020/grades/hw_2_table.csv')
d2[student_id == sid]

# HW 3
d3 <- fread('https://crossley.github.io/cogs2020/grades/hw_3_table.csv')
d3[student_id == sid]

# HW 4
d4 <- fread('https://crossley.github.io/cogs2020/grades/hw_4_table.csv')
d4[student_id == sid]

# HW 5
d5 <- fread('https://crossley.github.io/cogs2020/grades/hw_5_table.csv')
d5[student_id == sid]

# HW 6
d6 <- fread('https://crossley.github.io/cogs2020/grades/hw_6_table.csv')
d6[student_id == sid]

# HW 7
d7 <- fread('https://crossley.github.io/cogs2020/grades/hw_7_table.csv')
d6[student_id == sid]

# HW 8
d7 <- fread('https://crossley.github.io/cogs2020/grades/hw_8_table.csv')
d6[student_id == sid]

# HW 9
d7 <- fread('https://crossley.github.io/cogs2020/grades/hw_9_table.csv')
d6[student_id == sid]

# midterm
dm <- fread('https://crossley.github.io/cogs2020/grades/midterm.csv')
dm[student_id == sid]

# Midterm recovery 1
dm1 <- fread('https://crossley.github.io/cogs2020/grades/midterm_recovery_1.csv')
dm1[student_id == sid]

# Midterm recovery 2
dm2 <- fread('https://crossley.github.io/cogs2020/grades/midterm_recovery_2.csv')
dm2[student_id == sid]
```

## Total grades so far
```{r, eval=T, message=F, warning=F, fig.width=10}
# If you want to know where you are standing currently, run
# this entire code block and then select your student_id
# from dd
library(data.table)
library(ggplot2)

rm(list=ls())

d1 <- fread('https://crossley.github.io/cogs2020/grades/hw_1_table.csv')
d2 <- fread('https://crossley.github.io/cogs2020/grades/hw_2_table.csv')
d3 <- fread('https://crossley.github.io/cogs2020/grades/hw_3_table.csv')
d4 <- fread('https://crossley.github.io/cogs2020/grades/hw_4_table.csv')
d5 <- fread('https://crossley.github.io/cogs2020/grades/hw_5_table.csv')
d6 <- fread('https://crossley.github.io/cogs2020/grades/hw_6_table.csv')
d7 <- fread('https://crossley.github.io/cogs2020/grades/hw_7_table.csv')
d8 <- fread('https://crossley.github.io/cogs2020/grades/hw_8_table.csv')
d9 <- fread('https://crossley.github.io/cogs2020/grades/hw_9_table.csv')
dm <- fread('https://crossley.github.io/cogs2020/grades/midterm.csv')
dm1 <- fread('https://crossley.github.io/cogs2020/grades/midterm_recovery_1.csv')
dm2 <- fread('https://crossley.github.io/cogs2020/grades/midterm_recovery_2.csv')

d1[, assignment := 'hw1']
d2[, assignment := 'hw2']
d3[, assignment := 'hw3']
d4[, assignment := 'hw4']
d5[, assignment := 'hw5']
d6[, assignment := 'hw6']
d7[, assignment := 'hw7']
d8[, assignment := 'hw8']
d9[, assignment := 'hw9']
dm[, assignment := 'm']
dm1[, assignment := 'm1']
dm2[, assignment := 'm2']

d <- rbindlist(list(d1, d2, d3, d4, d5, d6, d7, d8, d9, dm, dm1, dm2), use.names=T, fill=T)
dd <- d[, .(score = unique(score)), .(student_id, assignment)]
m_subs_1 <- dm1[, unique(student_id)]
m_subs_2 <- dm2[, unique(student_id)]
dd <- dcast(dd, student_id ~ assignment, value.var = 'score')
dd[!(student_id %in% m_subs_1), m1 := m]
dd[!(student_id %in% m_subs_2), m2 := m]
dd[, hw1 := ifelse(is.na(hw1), 0, hw1)]
dd[, hw2 := ifelse(is.na(hw2), 0, hw2)]
dd[, hw3 := ifelse(is.na(hw3), 0, hw3)]
dd[, hw4 := ifelse(is.na(hw4), 0, hw4)]
dd[, hw5 := ifelse(is.na(hw5), 0, hw5)]
dd[, hw6 := ifelse(is.na(hw6), 0, hw6)]
dd[, hw7 := ifelse(is.na(hw7), 0, hw7)]
dd[, hw8 := ifelse(is.na(hw8), 0, hw8)]
dd[, hw9 := ifelse(is.na(hw9), 0, hw9)]
dd[, total_score := sum(0.1 * 1, # quizzes (assuming you got credit for the 3 given)
                    0.3 * mean(c(hw1, hw2, hw3, hw4, hw5, hw6, hw7, hw8, hw9)), # hw
                    0.3 * mean(c(m1, m2)), # midterm recovery
                    0.3 * 0.75, na.rm=F) # final (assuming 75%)
   , .(student_id)]
dd[, mark_class := "NA"]
dd[total_score >= .85, mark_class := "High Distinction"]
dd[total_score >= .75 & total_score < .85, mark_class := "Distinction"]
dd[total_score >= .65 & total_score < .75, mark_class := "Credit"]
dd[total_score >= .50 & total_score < .65, mark_class := "Pass"]
dd[total_score < .50, mark_class := "Fail"]
dd[, mark_class := factor(mark_class, 
                          levels=c('High Distinction',
                                   'Distinction', 
                                   'Credit', 
                                   'Pass', 
                                   'Fail'))]
ggplot(dd, aes(total_score, fill=mark_class)) +
  geom_histogram(bins=10, breaks=seq(0.0, 1.0, 0.05)) +
  xlim(0.3, 1.0)
```


## Homework 1 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_1_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')
```
  
## Homework 2 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_2_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')
```


## Homework 3 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_3_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')
```

## Homework 4 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_4_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Midterm 1 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/midterm_recovery_1.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)
```

## Midterm 2 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/midterm_recovery_2.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)
```

## Homework 5 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_5_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Homework 6 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_6_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Homework 7 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_7_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Homework 8 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_8_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10)

dd <- d[, .(ans_mean = mean(ans=="correct"),
            ans_err = sd(ans=="correct")/sqrt(.N)), .(ans_labels)]
ggplot(data=dd, aes(x=ans_labels, y=ans_mean)) +
  geom_pointrange(aes(x=ans_labels,
                      ymin=ans_mean-ans_err,
                      ymax=ans_mean+ans_err)) +
  xlab('Problem Number') +
  ylab('Proportion Correct')+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Homework 9 Grades
```{r, fig.width=10}
library(data.table)
library(ggplot2)

rm(list=ls())

d <- fread('https://crossley.github.io/cogs2020/grades/hw_9_table.csv')

dd <- d[, .(score = unique(score)), .(student_id)]
ggplot(dd, aes(score)) +
  geom_histogram(bins=10) +
  xlim(0, 1)
```