---
title: "Homework 2"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(data.table)
library(ggplot2)
```

<style type="text/css">
.table {

    width: 40%;

}
</style>

<!-- ##### Special Note -->
<!-- Let  -->

<!-- $Y \sim a X + b$ -->

<!-- Then: -->

<!-- \begin{align} -->
<!-- E(Y) &= E(a X + b) \\ -->
<!--      &= E(a X) + E(b) \\ -->
<!--      &= a E(X) + b \\ -->
<!-- \\ -->
<!-- \\ -->
<!-- Var(Y) &= Var(a X + b) \\ -->
<!--        &= Var(a X) + Var(b) \\ -->
<!--        &= a^2 Var(X) \\ -->
<!-- \end{align} -->


## Problem 1

Consider an experiment in which two four-sided die are
thrown. 

**(a)** Create a `data.table` named `ans_1a` with two columns
labelled `die1` and `die2` where these columns encode the
**sample space** of this experiment. See below for the first
eight rows of the resulting `data.table`.

```{r echo=FALSE}
die1 <- rep(c(1, 2, 3, 4), 4)
die2 <- rep(c(1, 2, 3, 4), each=4)
ans_1a <- data.table(die1, die2)
ans_1a[1:8]
```

**(b)** Create a deep copy of `ans_1a` called `ans_1b` and
then add a column to `ans_1b` called `X` and set it equal to
the sum of the corresponding `die1` and `die2` values on a
per row basis.

**(c)** Let $X$ be the random variable defined by the `X`
column in `ans_1b`. What is $P(X<5)$? Store your result in a
variable named `ans_1c`.

**(d)** Is the event defined as the set of all outcomes for
which $X=2$ an elementary event? Store your answer (`"YES"`
or `"NO"`) in a variable named `ans_1d`.

## Problem 2

For the following probability distribution:

<div width:50px>
| $x$    |  2 |  3 |  4 |  5 |  6 |
|--------|----|----|----|----|----|
| $f(x)$ | .1 | .3 | .3 | .2 | .1 |
</div>

**(a)** Is $f(x)$ a valid probability distribution? Store
your answer (`"YES"` or `"NO"`) in a variable named
`ans_2a`.

**(b)** Calculate $E(X)$ store the result in a variable
named `ans_2b`.

**(c)** Calculate $P(X \geq 4)$ store the result in a
variable named `ans_2c`.

**(d)** Calculate $P(2 < X \leq 4)$ and store the result in
a variable named `ans_2d`.

<!-- **(d)** Let $Y = (2X-8)^2$.  Obtain the distribution of $Y$ and calculate $E(Y)$ -->

## Problem 3

Given the following probability distribution:

```{r, echo=FALSE}
x <- c(0, 1, 2, 3)
fx <- c(0.3, 0.4, 0.2, 0.1)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) + 
  geom_point() +
  geom_segment(aes(x=x, y=0, xend=x, yend=fx)) +
  xlab("x") +
  ylab("P(X=x)")
```

Find:

**(a)** Calculate $P(X \geq 2)$ and store your result in a
variable named `ans_3a`.

**(b)** Calculate $P(0 < X \leq 2)$ and store your result in
a variable named `ans_3b`.

**(c)** Calculate $E(X)$ and store your result in a variable
named `ans_3c`.

**(d)** $Var(X)$ and store your result in a variable named
`ans_3d`.

**(e)** $sd(X)$ and store your result in a variable named
`ans_3e`.

<!-- ## Problem 3 -->

<!-- For scores $X$ on a nationally administered aptitude test, the mean and variance are: -->

<!-- $E(X)=120$ -->

<!-- $Var(X)=100$ -->

<!-- Find the mean and variance of: -->

<!-- **(a)** $Y = (X-120)/10$ -->

<!-- **(b)** $Y = (X-100)/20$ -->

## Problem 4

The chance that monkeys will have a positive (*successful*)
reaction during a trial of an experiment is $1/3$. Trials
are performed until the first success occurs. Define the
following random variable $X$:

$X$ = number of failures before the first success.

What is the probability distribution of $X$?

Let $S$ denote success and $F$ denote failure Then the
sample space is: $S = {S, FS, FFS, FFFS, ...}$

You can see that it's impossible to make a complete list of
all the values and their probabilities. However, this is a
well studied problem, and it turns out that there is an
formula that will give $P(X=x)$ for any $x$: 

$$P(X=x) = p(1-p)^x$$

This is called the geometric distribution, and it is defined
by a single parameter $p$ (the probability of success). In
`R`, you can use the function `dgeom()` and `pgeom()` to
obtain geometric probabilities.

Let $P(Success) = 1/3$.

**(a)** Is the geometric distribution continuous or
discrete. Use either `ans_4a<-"continuous"` or
`ans_4a<-"discrete"` to store your answer.

**(b)** Plot the geometric probability distribution from
$x=0$ to $x=10$ and $p=\frac{1}{3}$. To do this, perform the
following steps:

* Define a variable `p` and set its value to `1/3`

* Define a variable `x` and set its value to `0:10`

* Define a variable `fx` and set its value to `dgeom(x, p)`

* Define a variable `d` and set its values to
`data.table(x,fx)`

* Use `ggplot`, `geom_point` and `geom_segment` to replicate
the following plot exactly and store your result in a
variable named `ans_4b`.

```{r, echo=FALSE}
p <- 1/3
x <- 0:10
fx <- dgeom(x, p)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_point() +
  geom_segment(aes(x=x, xend=x, y=0, yend=fx))
```

<!-- **(c)** Plot the geometric distribution for the following -->
<!-- values of $P(Success)$: -->

<!-- $$ -->
<!-- p(Success) = 0.2 \\ -->
<!-- p(Success) = 0.5 \\ -->
<!-- p(Success) = 0.8 \\ -->
<!-- $$ -->

<!-- Use `facet_wrap` to plot on a single figure. -->

**(c)** Compute the expected value of the geometric
distribution that you plotted in (b) using
$\sum_{i=0}^{10}xP(x)$. Store your result in a variable
named `ans_4c`.

**(d)** It turns out it can be proven that if $X$ ~
geometric with probability of success of $p$, then
$E(X)=\frac{1-p}{p}$. This formula will yield a slightly
different number than the number you computed in c. Why is
this case?

* Use `ans_4d <- "option 1"` if the difference arises due to
computer rounding errors.

* Use `ans_4d <- "option 2"` if the difference arises
because part c only considered x in the range 0 to 10, yet
the true range of a geometric random variable is 0 to
infinity.

* Use `ans_4d <- "option 3"` if the difference arises
because the distribution of sample means is a random
variable.


<!-- ## Probalem 4 -->

<!-- **(a)** Compute $P(X = 3)$ and store your result in a -->
<!-- variable named `ans_2a`. -->

<!-- **(b)** Compute $P(X > 3)$ and store your result in a -->
<!-- variable named `ans_2b`. -->

<!-- **(c)** Compute $P(X >= 3)$ and store your result in a -->
<!-- variable named `ans_2c`. -->

<!-- **(d)** Compute $P(X <= 3)$ and store your result in a -->
<!-- variable named `ans_2d`. -->

## Problem 5

**(a)** Does the following histogram represent samples from
a geometric distribution (use `ans_5a <- "geometric"`) or
from the distribution of geometric sample means `ans_5a <-
"geometric means"`? Note that histograms can make samples
from a discrete distribution appear continuous.

<!-- **(c)** Draw a sample of size $n=1000$ from the above -->
<!-- geometric distribution and plot it with a histogram. To do -->
<!-- this, perform the following steps exactly: -->

<!-- * Define a variable `n` and set its value to `1000` -->

<!-- * Define a variable `x` and set its value to `rgeom(n, p)` -->

<!-- * Define a variable `d` and set its value to `data.table(x)` -->

<!-- * Use `ggplot`, `geom_histogram(bins=10)`to replicate the -->
<!-- following plot (your bars will be different heights because -->
<!-- `rgeom` gives a random sample): -->

```{r, echo=FALSE}
n <- 1000
x <- rgeom(n, p)
d <- data.table(x)
ggplot(d, aes(x)) + 
  geom_histogram(breaks=0.5:10.5)
```


**(b)** Does the following histogram represent samples from
a geometric distribution (use `ans_5b <- "geometric"`) or
from the distribution of geometric sample means `ans_5b <-
"geometric means"`? Note that histograms can make samples
from a discrete distribution appear continuous.

```{r, echo=FALSE}
n <- 1000
n_exp <- 100
d_list <- vector("list", n_exp)
for(i in 1:n_exp) {
  x <- rgeom(n, p)
  d <- data.table(x = mean(x))
  d_list[[i]] = d
}
dd <- rbindlist(d_list)
ggplot(dd, aes(x)) + 
  geom_histogram(bins=10) 
```


<!-- **(g)** Plot samples from a geometric distribution for the -->
<!-- following values of -->

<!-- $$ -->
<!-- p(Success) = 0.2 \\ -->
<!-- p(Success) = 0.5 \\ -->
<!-- p(Success) = 0.8 \\ -->
<!-- $$ -->

<!-- Use facet_wrap to plot on a single figure. -->

<!-- **(h)** For each panel in (f), draw a vertical line -->
<!-- indicating the sample mean -->

<!-- **(i)** Repeat (g) and (h) a few times to see the variation -->
<!-- in sample mean, and the absence of variance in the -->
<!-- population mean. Make a figure that shows the results of -->
<!-- each of these experiments in a separate panel of one figure. -->

<!-- **(j)** Make a histogram to estimate the distribution of -->
<!-- sample means -->


<!-- ## Problem 5 -->

<!-- **(a)** Read the following csv file into a `data.table` -->

<!-- https://crossley.github.io/cogs2020/data/rb_bs_delay/rbbsdelay.csv -->

<!-- `condition`, `subject` and `block` are **independent variables**. This means -->
<!-- that they are controlled by the experimenter. `acc_mean` and `rt_mean` are -->
<!-- **dependent variables**. This means that they are determined by the random -->
<!-- samples taken as part of performing an experiment. -->

<!-- This is data from a fairly simple category learning experiment where I was again -->
<!-- investigating the effect of feedback delay on learning, but this time I took a -->
<!-- different approach than the trials to criterion approach we worked with in -->
<!-- previous lectures and homework. -->

<!-- **(b)** For each column in the resulting `data.table`, say whether the data is -->
<!-- continuous, discrete, or categorical. -->

<!-- **(c)** Is this a within-subject or a between-subject design? Answer this by -->
<!-- probing the `data.table`. If it within-subject, then the same subjects will be -->
<!-- repeated in different conditions and blocks. -->

<!-- **(d)** Plot the *learning curves* for this experiment: That is, plot mean -->
<!-- accuracy per block and per condition across subjects on the y-axis and plot -->
<!-- block on the x-axis. Include error bars indicating the standard error. -->

<!-- **(e)** It is common to see speed-accuracy trade-offs in behavioural data of all -->
<!-- sorts. That is, it is common for participants to be less accurate when they -->
<!-- decide quickly, and more accurate when the decide slowly. Use `acc_mean` -->
<!-- and`rt_mean` to look for evidence of a speed-accuracy trade-off in this data. Do -->
<!-- you see any? NOTE: just make a plot and write a sentence. Nothing fancy here. -->

