---
title: "Homework 7"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r, echo=F, message=F, warning=F}
library(data.table)
library(ggplot2)
library(ggpubr)
rm(list = ls())

gen_exp <- function(muA1, muA2, muB1, muB2, muC1, muC2, sig, n) {

  ## generate obervations
  y_A1 <- rnorm(n, muA1, sig)
  y_A2 <- rnorm(n, muA2, sig)
  y_B1 <- rnorm(n, muB1, sig)
  y_B2 <- rnorm(n, muB2, sig)
  y_C1 <- rnorm(n, muC1, sig) # new level but same mu
  y_C2 <- rnorm(n, muC2, sig) # new level but same mu

  ## store observations plus relevant indicators in data.table
  d <- data.table(y=c(y_A1, y_A2, y_B1, y_B2, y_C1, y_C2),
                  treatment=rep(c('A', 'B', 'C'), each=2*n),
                  dose=rep(c(1, 2), 3, each=n))

  d[, treatment := factor(treatment)]
  d[, dose := factor(dose)]

  ## main effect of treatment (A vs B)
  dd <- d[, .(mean(y), sd(y)/sqrt(length(unique(dose)))), .(treatment)]
  g1 <- ggplot(dd, aes(treatment, V1)) +
    geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
    theme_classic() +
    theme(aspect.ratio=1) +
    ylab('Observed mean effect') +
    ggtitle('Main effect')

  ## main effect of dose (1 vs 2)
  dd <- d[, .(mean(y), sd(y)/sqrt(length(unique(treatment)))), .(dose)]
  g2 <- ggplot(dd, aes(dose, V1)) +
    geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
    theme_classic() +
    theme(aspect.ratio=1) +
    ylab('Observed mean effect') +
    ggtitle('Main effect')

  ## interaction between treatment and dose
  dd <- d[, .(mean(y), sd(y)/sqrt(length(unique(treatment))*length(unique(dose)))), .(treatment, dose)]
  g3 <- ggplot(dd, aes(dose, V1, colour=treatment)) +
    geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
    geom_line(aes(as.integer(dose), V1, colour=treatment)) +
    theme_classic() +
    theme(aspect.ratio=1) +
    ylab('Observed mean effect') +
    ggtitle('Interaction')

  return(list(g1, g2, g3, unique(d)))
}
```

**1.** Considering the following data and using a two-way
ANOVA, test the hypothesis that there is a main effect of
treatment, a main effect of dose, or an interaction between
treatment and dose. Please populate the following variables.

```{r, echo=F}
set.seed(0)
n <- 4
res <- gen_exp(10, 30, 40, 20, 40, 20, 5, n)
g1 <- res[[1]]
g2 <- res[[2]]
g3 <- res[[3]]
d <- res[[4]]
d
```

```{r, eval=F}
# Please compute the following by hand as demonstrated in the lecture notes

# Values corresponding to the main effect of treatment
prob_1_theta_hat_obs_treatment <- theta_hat_obs_treatment # observed value of test statistic
prob_1_pval_treatment <- p_val_treatment # p-value of observed test statistic
prob_1_theta_hat_crit_treatment <- theta_hat_crit_treatment # critical value of test statistic

# Values corresponding to the main effect of dose 
prob_1_theta_hat_obs_dose <- theta_hat_obs_dose
prob_1_pval_dose <- p_val_dose
prob_1_theta_hat_crit_dose <- theta_hat_crit_dose

# Values corresponding to the interaction of treatment and dose
prob_1_theta_hat_obs_interaction <- theta_hat_obs_interaction
prob_1_pval_interaction <- p_val_interaction
prob_1_theta_hat_crit_interaction <- theta_hat_crit_interaction

# the result of ezANOVA() containing information about both
# main effects and their interaction.
prob_1_r_result <- r_result 
```

**2.** Consider the following results

```{r, echo=F, fig.width=10}
ggarrange(g1, g2, g3, nrow=1, widths=c(1, 1, 1.3))
```

* What effects do you expect to be significant, or close to significant?
  * `ans_2a <- "Dose, treatment, and interaction"` 
  * `ans_2a <- "Dose and interaction"` 
  * `ans_2a <- "Treatment and interaction"` 
  * `ans_2a <- "Dose only"` 
  * `ans_2a <- "Treatment only"` 
  * `ans_2a <- "Interaction only"` 

**3.** Consider the data located here:

https://crossley.github.io/cogs2020/data/maze_2/maze_2.csv

This data contains the results of an experiment in which
researchers examined the effect of maze odour and maze
material on maze run time in a large sample of rats.

* Every row corresponds to a different rat.

* `y` contains the amount of time in seconds taken to
complete the maze.

* `maze_odour` indicates the odour of the maze.

* `maze_material` indicates the material of maze.

Using two-way ANOVA, assess any significant differences
between levels of `maze_odour`, `maze_material`, or their
interactions?

```{r, eval=F}
# You do not have to perform the two-way ANOVA steps
# manually. Please use ezANOVA() and store the result of
# this function to a variable named `ans_3`
ans_3 <- ezANOVA(...)
```

**4.** How many t-tests would you have to perform in order
to test all the hypotheses tested in problem 3?

```{r, eval=F}
# Your response to this question should be an integer
ans_4 <- 'put_your_integer_answer_here'
```