---
title: "Homework 3"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

<div style="color:#990000">
**Note that this homework is not yet finalised. I am posting
it here for practice purposes only.**
<br></br>
</div>

<style type="text/css">
.table {

    width: 40%;

***
</style>


```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)
```


## Problem 1

**(a)** Verify that $p(X<x)=p(X \leq x)$ for:

  * $X\sim\mathcal{N}(\mu,\sigma)$ 
  * $x=1$ 
  * $\mu=0$ 
  * $\sigma=1$ 
  
  * Do this by performing the following steps:

  * Define a variable `p1` and set its value to $p(X<x)$.
  Use `lower.tail=FALSE` to do so.
  
  * Define a variable `p2` and set its value to
  $p(X{\leq}x$. Use `lower.tail=TRUE` to do so.
  
  * Define a variable `ans_1a` and set its value to `TRUE`
  if `p1` is equal to `p2` and set its value to `FALSE`
  otherwise.
  
  * The $X$ probability distribution (pdf since X is
  continuous) and the probabilities given by
  `lower.tail=TRUE` and `lower.tail=FALSE` are illustrated
  in the following figure. This figure is only here to help
  you think about what is being asked.

```{r, echo=F}
x <- seq(-5, 5, 0.01)
fx <- dnorm(x)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_line() +
  geom_segment(aes(x=1, xend=1, y=0, yend=dnorm(1)), linetype=2) +
  geom_ribbon(data=d[x<=1],
              aes(x=x, ymin=0, ymax=fx),
              fill='blue',
              alpha=0.25) +
  geom_ribbon(data=d[x>1],
              aes(x=x, ymin=0, ymax=fx),
              fill='green',
              alpha=0.25) +
  annotate("text", x=-0.5, y=0.05, label="P(X<=x)") +
  annotate("text", x=1.5, y=0.05, label="P(X>x)") +
  ylab('f(x)')
```


**(b)** Verify that $p(X<x){\neq}p(X{\leq}x)$ for:

  * $X{\sim}binom(n,p)$ 
  
  * $x=3$
  
  * $n=10$
  
  * $p=0.2$. 

  * Do this by performing the following steps:

  * Define a variable `p1` and set its value to $p(X<x)$.
  Use `lower.tail=FALSE` to do so.
  
  * Define a variable `p2` and set its value to
  $p(X{\leq}x$. Use `lower.tail=TRUE` to do so.
  
  * Define a variable `ans_1b` and set its value to `TRUE`
  if `p1` is equal to `p2` and set its value to `FALSE`
  otherwise.
  
  * The $X$ probability distribution (pdf since X is
  continuous) and the probabilities given by
  `lower.tail=TRUE` and `lower.tail=FALSE` are illustrated
  in the following figure. This figure is only here to help
  you think about what is being asked.

```{r, echo=F}
n <- 10
p <- 0.2
x <- seq(0, n, 1)
fx <- dbinom(x, n, p)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_segment(data=d[x<=3], aes(x=x, xend=x, y=0, yend=dbinom(x,n,p)), colour='blue') +
  geom_segment(data=d[x>3], aes(x=x, xend=x, y=0, yend=dbinom(x,n,p)), colour='green') +
  annotate("text", x=3, y=0.21, label="P(X<=x)") +
  annotate("text", x=4, y=0.1, label="P(X>x)") +
  scale_x_continuous(breaks=0:10) +
  ylab('f(x)')
```

## Problem 2

Consider the following random variables:

* $X{\sim}\mathcal{N}(\mu_X=15,\sigma_X=2.74)$ 

* $Y{\sim}Binomial(n=30,p=0.5)$.

The $X$ and $Y$ distributions are illustrated in the
following figure:

```{r, echo=F}
x <- seq(0, 30, 0.01)
fx <- dnorm(x, 15, 2.74)
d <- data.table(x, fx)
gx <- ggplot(d, aes(x, fx)) +
  geom_line() +
  ylab('f(x)')

n <- 30
p <- 0.5
y <- seq(0, 30, 1)
fy <- dbinom(y, n, p)
d <- data.table(y, fy)
gy <- ggplot(d, aes(y, fy)) +
  geom_segment(aes(x=y, xend=y, y=0, yend=fy)) + 
  ylab('f(y)')

ggarrange(gx, gy, ncol=2)
```

**(a)** Compute $P(X > 17)$ and store it in a variable
named `ans_2a`.

**(b)** Compute $P(X \leq 14)$ and store it in a variable
named `ans_2b`.

**(c)** Compute $P(X \geq 20)$ and store it in a variable
named `ans_2c`.

**(d)** Compute $P(X < 13)$ and store it in a variable
named `ans_2d`.

**(e)** Compute $P(Y > 17)$ and store it in a variable
named `ans_2e`.

**(f)** Compute $P(Y \leq 14)$ and store it in a variable
named `ans_2f`.

**(g)** Compute $P(Y \geq 20)$ and store it in a variable
named `ans_2g`.

**(h)** Compute $P(Y < 13)$ and store it in a variable
named `ans_2h`.

**(i)** How do all of the above quantities computed with
respect to $X$ compare to those computed with respect to
$Y$? No points for this one. Just think about it.

<!-- **(e)** Compute $P(12.22 < X < 24.55)$ and store it in a -->
<!-- variable named `ans_2e`. -->

<!-- **(f)** Compute $x$ such that $P(X>x) = .9$ and store it in -->
<!-- a variable named `ans_2f`. -->

<!-- **(g)** Compute $x$ such that $P(X \leq x) = .8$ and store -->
<!-- it in a variable named `ans_2g`. -->

## Problem 3

**(a)** Which of the following plots corresponds to the
quantity requested in problem 2a? Please respond by creating
a variable named `ans_3a` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(b)** Which of the following plots corresponds to the
quantity requested in problem 2b? Please respond by creating
a variable named `ans_3b` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(c)** Which of the following plots corresponds to the
quantity requested in problem 2c? Please respond by creating
a variable named `ans_3c` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2d? Please respond by creating
a variable named `ans_3b` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2e? Please respond by creating
a variable named `ans_3e` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2f? Please respond by creating
a variable named `ans_3f` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2g? Please respond by creating
a variable named `ans_3g` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2h? Please respond by creating
a variable named `ans_3h` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`.

```{r, echo=F}
xcritvec <- c(17, 14, 20, 13)

mu <- 15
sigma <- 2.74

x <- seq((mu-sigma*4), (mu+sigma*4), 0.01)
fxd <- dnorm(x, mu, sigma)

fxp <- pnorm(x, mu, sigma, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[1]
ga1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x>xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

ga2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  theme(aspect.ratio=1)

fxp <- pnorm(x, mu, sigma, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[2]
gb1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x <= xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

gb2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=T),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=T),
               yend=pnorm(xcrit, mu, sigma, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  theme(aspect.ratio=1)

fxp <- pnorm(x, mu, sigma, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[3]
gc1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x >= xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

gc2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  theme(aspect.ratio=1)

fxp <- pnorm(x, mu, sigma, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[4]
gd1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x < xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

gd2n <- ggplot(d, aes(x, 1-fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X<=x)') +
  theme(aspect.ratio=1)
```


```{r, echo=F}
n <- 30
p <- 0.5
x <- seq(0, n, 1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[1]
ga1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x>xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x<=xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))+
  theme(aspect.ratio=1)

ga2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit, 
               xend=xcrit, 
               y=0, 
               yend=pbinom(xcrit, n, p, lower.tail=F), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pbinom(xcrit, n, p, lower.tail=F),
               yend=pbinom(xcrit, n, p, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[2]
gb1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x<=xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x>xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

gb2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit, 
               xend=xcrit, 
               y=0, 
               yend=pbinom(xcrit, n, p, lower.tail=T), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pbinom(xcrit, n, p, lower.tail=T),
               yend=pbinom(xcrit, n, p, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[3]
gc1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x>=xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x<xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))+
  theme(aspect.ratio=1)

gc2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit-1, 
               xend=xcrit-1, 
               y=0, 
               yend=pbinom(xcrit-1, n, p, lower.tail=F), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit-1,
               y=pbinom(xcrit-1, n, p, lower.tail=F),
               yend=pbinom(xcrit-1, n, p, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[4]
gd1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x<xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x>=xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

gd2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit-1, 
               xend=xcrit-1, 
               y=0, 
               yend=pbinom(xcrit-1, n, p, lower.tail=T), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit-1,
               y=pbinom(xcrit-1, n, p, lower.tail=T),
               yend=pbinom(xcrit-1, n, p, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)
```

```{r, echo=F, fig.height=25}
plot_list <- list(ga1n, gb1n, gc1n, gd1n,
                  ga2n, gb2n, gc2n, gd2n,
                  ga1b, gb1b, gc1b, gd1b,
                  ga2b, gb2b, gc2b, gd2b)

set.seed(0)
plot_list <- sample(plot_list, 8)

plot_labels <- list('I', 'II', 'III', 'IV',
                    'V', 'VI', 'VII', 'VIII')

ggarrange(plotlist=plot_list,
          labels=plot_labels,
          nrow=8,
          ncol=2)
```


## Problem 4

Consider the random variable $X{\sim}F(d1,d2)$ for $d1=5$,
$d2=100$.  For these values of $d1$ and $d2$,

$E[X] = \frac{d2}{d2 - 2}$

$Var[X] = \frac{2 d2^2 (d1 + d2 - 2)}{d1 (d2 - 2)^2 (d2 - 4)}$

```{r, echo=F}
d1 <- 5
d2 <- 100
```


**(a)** Compute the population mean and variance for $X$ ans
store your result in a variable named `ans_4a`.

**(b)** We next perform 6 experiments. In each experiment,
we draw 1000 samples from $X$ and plot a histogram for the
results of each experiment. Please copy and paste the the
code below into your homework .R file and then run the code.

```{r}
# Define the number of experiments to run
n_exp <- 6

# Define the number of samples to draw per experiment
n <- 1000

# Create an empty list. We will later use this list to store
# the results of each experiment
storage_list <- list()

# Run through a for loop and in each iteration of the for
# loop, draw n samples from the random variable X, and store
# the result in the storage list defined just above.
for(i in 1:n_exp) {
  
  # We will use the function `rf()` to draw random samples
  # from X. However, I need you all to work with the same
  # random samples as me (for grading purposes), so we use
  # the `set.seed()` function to ensure that we all get the
  # same "random" numbers.
  set.seed(i)
  
  # Generate a random sample using the `rf()` function
  random_sample <- rf(n, d1, d2)
  
  # we will want things in data.table format later, so might
  # as well get started on it now.
  result <- data.table(exp=i, x=random_sample)
  
  # Store the result in our storage list
  storage_list[[i]] <- result
}

# Storage_list contains a list of data.tables. Each
# data.table in this list has one column named exp (this is
# an indicator column that indicates the experiment from
# which data was obtained) and another named x (this column
# contains the data sampled in each experiment). We next use
# the data.table function `rbindlist()` to combine this list
# of data.tables into a single data.table.
d <- rbindlist(storage_list)

# We now plot histograms to show the results of each
# experiment.
ggplot(d, aes(x)) +
  geom_histogram(bins=50) +
  facet_wrap(~exp)
```

Do the histograms appear Normally distributed? If not, what
is the main difference? Please respond with one of the
following:

* `ans_4b <- "skewed left"`
* `ans_4b <- "skewed right"`
* `ans_4b <- "Normal"`

**(c)** We next estimate the distribution of sample means as
a function of the number of samples `n` that were drawn per
experiment. We will perform 6 experiments in which we draw
2, 10, 20, 30, 50, and 100 samples from $X$ (each of these
numbers corresponds to one experiment). Each of these
experiments, if performed once, will give us one sample
mean. To examine the distribution of sample means, we repeat
each experiment 1000 times. This will give us 1000 sample
means per experiment. For each experiment, we estimate the
distribution of sample means with a histogram and make a
single figure with separate panels for each experiment to
show our results.  Please copy and paste the the code below
into your homework .R file and then run the code.

```{r}
# The following code is very similar to the above, so no
# comments are provided.
n_exp <- 1000
n <- c(2, 5, 10, 15, 30, 50)

storage_list <- list()
for(j in 1:length(n)) {
  for(i in 1:n_exp) {
    set.seed(i + j*n_exp)
    random_sample <- rf(n[j], d1, d2)
    sample_mean <- mean(random_sample)
    result <- data.table(exp=i, n=n[j], x=sample_mean)
    storage_list[[i + j*n_exp]] <- result
  }
}

d <- rbindlist(storage_list)

ggplot(d, aes(x)) +
  geom_histogram(bins=50) +
  facet_wrap(~n)
```

* What happens to the variance of the distribution of sample
means as $n$ increases?
  * `ans_4c1 <- "Stays the same"`
  * `ans_4c1 <- "Increases"`
  * `ans_4c1 <- "Decreases"`

* What happens to the skew of the distribution of sample
means as $n$ increases?
  * `ans_4c2 <- "Stays the same"`
  * `ans_4c2 <- "Increases"`
  * `ans_4c2 <- "Decreases"`

**(d)** What is the population mean of this distribution of
sample means? Store your answer in a variable named
`ans_4d`.

**(e)** What is the population variance of this distribution
of sample means? Store your answer in a variable named
`ans_4e`.

**(f)** How does increasing $n$ influence your sample mean
estimate of the population mean?

  * `ans_4f <- "estimate approaches the population value"`
  * `ans_4f <- "no effect"`
  * `ans_4f <- "population value decreases variance"`

