---
title: "Homework 6 Solutions"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Solutions

```{r}
# Basic setup code that you will need for this problem
library(data.table)
library(ggplot2)

rm(list=ls())

# basic id info
my_name <- "John Doe"
my_student_id <- "12345678"

# Please include the following line of code as well
set.seed(0)

# 1
# Assume X ~ Normal(mu_x, sig_x)
# Then Null and alternative H's are:
# H0: mu_x = 58
# H1: mu_x != 58

# type I error (alpha) = 0.05
type_I_error <- 0.05

# estimate mu_x with mu_x_hat = x_bar
x <- c(63.97,
      60.29,
      85.60,
      72.57,
      54.65,
      53.74,
      69.13,
      70.07,
      67.19,
      53.95
     )

n <- length(x)
x_bar_obs <- mean(x)

# define the sampling distribution of x_bar:
#
# X ~ Normal(mu_x, sig_x)
# X_bar ~ Normal(mu_x_bar, sig_x_bar)
#
# mu_x_bar = mu_x
# sig_x_bar = sig_x / sqrt(n)

mu_x = 58
sig_x = sqrt(10) # given in the problem

mu_x_bar <- mu_x # from central limit theorem
sig_x_bar <- sig_x / sqrt(n) # from central limit theorem

# prepare for two-tailed calculations
x_bar_obs_lower <- mu_x_bar - abs(x_bar_obs - mu_x_bar)
x_bar_obs_upper <- mu_x_bar + abs(x_bar_obs - mu_x_bar)

# critical value
critical_value_lower <- qnorm(type_I_error, mu_x_bar, sig_x_bar, lower.tail=T)
critical_value_upper <- qnorm(type_I_error, mu_x_bar, sig_x_bar, lower.tail=F)

# CI
w <- critical_value_upper - critical_value_lower
CI_lower <- x_bar_obs - w/2
CI_upper <- x_bar_obs + w/2

# p-value
p_value_lower <- pnorm(x_bar_obs_lower, mu_x_bar, sig_x_bar, lower.tail=T)
p_value_upper <- pnorm(x_bar_obs_upper, mu_x_bar, sig_x_bar, lower.tail=F)

# Populate marking variables
ans_1_test_stat_obs <- x_bar_obs
ans_1_critical_value_lower <- critical_value_lower
ans_1_critical_value_upper <- critical_value_upper
ans_1_CI_lower <- CI_lower
ans_1_CI_upper <- CI_upper
ans_1_p_value <- p_value_lower + p_value_upper

# 2
# Assume X ~ Binomial(n, p)
# Then Null and alternative H's are:
# H0: p=0.5
# H1: p!=0.5

# type I error (alpha) = 0.05
type_I_error <- 0.05

# estimate p with the observed proportion of successes
n <- 100 # given in problem
n_obs = 63 # given in the problem
p_hat_obs = n_obs / n

# define the sampling distribution of np:
# np ~ Binomial(n, p)
p <- 0.5 # as specified in H0

# prepare for two-tailed calculations
n_obs_lower <- p * n - abs(n_obs - p * n)
n_obs_upper <- p * n + abs(n_obs - p * n)

# critical value (note the -1 due to discrete distribution)
# NOTE: You guys should ask Shir about the +1/-1
critical_value_lower <- (qbinom(type_I_error/2, n, p, lower.tail=T) - 1) / n
critical_value_upper <- (qbinom(type_I_error/2, n, p, lower.tail=F) + 1) / n

# CI
w <- critical_value_upper - critical_value_lower
CI_lower <- p_hat_obs - w/2
CI_upper <- p_hat_obs + w/2

# p-value
p_value_lower <- pbinom(n_obs_lower, n, p, lower.tail=T)
p_value_upper <- pbinom(n_obs_upper - 1, n, p, lower.tail = F)

# Populate marking variables
ans_2_test_stat_obs <- p_hat_obs
ans_2_critical_value_lower <- critical_value_lower
ans_2_critical_value_upper <- critical_value_upper
ans_2_CI_lower <- CI_lower
ans_2_CI_upper <- CI_upper
ans_2_p_value <- p_value_lower + p_value_upper

# 3
ans_3_critical_value_lower <- 0.025
ans_3_critical_value_upper <- 1 - 0.025
w <- ans_3_critical_value_upper - ans_3_critical_value_lower
ans_3_CI_lower <- 0.75 - w/2
ans_3_CI_upper <- 0.75 + w/2
ans_3_p_value <- 0.5

# 4
ans_4a <- "confidence"
ans_4b <- "beta"
ans_4c <- "alpha"
ans_4d <- "power"
ans_4e <- "alpha"
ans_4f <- "power"