---
title: "Homework 5"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 18pt
---
```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
```

```{r setup, include=FALSE, message=F}
knitr::opts_chunk$set(echo = T, collapse=T, warning=F)
```

<style type="text/css">
  body{
  font-size: 18pt;
}

<style type="text/css">
.table {

    width: 40%;
}
</style>

```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)
library(latex2exp)
```

## 0.
* Create a new file named `cogs2020_hw_5.R` and save it in
the `cogs2020` folder that you created during Lecture 1.

* If you have not already installed the `data.table` and the
`ggplot2` packages, do so now. If you use the
`install.packages` function, do not include it in the script
that you hand in.

* Make the first line of your script `library(data.table)`
and the second line `library(ggplot2)`.

* Make the second line of your script `rm(list=ls())`. This
line of code will erase any variable defined before it. Do
not put this line of code after anything important that you
want to keep (e.g., so that I can mark your work).

* After `rm(list=ls())`, create a variable named `my_name`
and set its value equal to a **character** vector (i.e.,
letters surrounded by `""` or by `''`) containing your name.

* After `rm(list=ls())`, create a variable named
`my_student_id` and set its value equal to a **character**
vector (i.e., letters surrounded by `""` or by `''`)
containing your student id.

* So far, here is what your file should look like if your
name is `John Doe` and and your student ID is `12345678`.
Note that the following code chunk also includes a few lines
to create a `data.table` named `d` that many of the problems
below require you manipulate.
```{r, echo=T}
# load the packages we will need
library(data.table)
library(ggplot2)

# clean session
rm(list=ls())

# basic id info
my_name <- "John Doe"
my_student_id <- "12345678"

# Please include the following line of code as well
set.seed(0)
```

## 1.
Suppose that researchers for a popular social media platform
are interested maximising the amount of time that users
spend on their app, and that they are faced with a tough
decision about whether or not the app should use a
continuous scroll feature. Since opinions on the research
team are split on the issue, they decide to perform an
experiment to clear things up. They implement the continuous
scroll feature and push it to their users. The average time
that a user spends on app currently is 58 minutes per day.
Thus they are interested in whether or not continuous scroll
increases this number. Suppose they obtain the following
sample from this experiment:

$$
x = \{ 63.97, 60.29, 85.60, 72.57, 54.65, 53.74, 69.13, 70.07, 67.19, 53.95 \}
$$

Test the hypothesis that infinite scroll increased time on
app beyond 58 minutes per day assuming that the population
variance of $X$ is $\sigma_X^2 = 10$. **Do not use `t.test`,
`binim.test` or any other built in full test function unless
explicitly instructed to do so.**

* Store the value of your observed test statistic in a
variable named `ans_1_test_stat_obs`.

* Store the p-value of your test in a variable named
`ans_1_p_value`.


## 2.
Suppose that a cognitive neuroscientist would like to
discover if the response a participant makes in a
2-alternative forced choice task can be predicted on the
basis of neural activity occurring 1 second before the motor
response is made. They train a machine learning classifier
to do this, and discover that it correctly predicted
participant choices in 63 out of 100 trials. Test the
hypothesis that their machine learning classifier predicted
participant choice at above-chance levels. **Do not use
`t.test`, `binim.test` or any other built in full test
function unless explicitly instructed to do so.**

* Store the value of your observed test statistic in a
variable named `ans_2_test_stat_obs`.

* Store the p-value of your test in a variable named
`ans_2_p_value`.


## 3.
Test the hypothesis that $\theta$ is greater than 6, where
$\theta$ is a parameter of some random variable.  The
sampling distribution for $\hat{\theta}$ is shown in the
following figure and the observed test statistic
$\hat{\theta}_{obs}$ is shown in blue.  Ensure that the type
I error rate of your test is at least $\alpha=0.05$.

```{r, echo=F, include=T}
n <- 6
p <- 0.4
x <- seq(0, n, 1)
fx <- dbinom(x, n, p)
x <- seq(0, 2*n, 2)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_point() +
  geom_segment(aes(x=8, y=0, xend=8, yend=fx[5]), colour='blue') +
  scale_x_continuous(breaks=x) +
  scale_y_continuous(breaks=fx) +
  xlab('x') +
  ylab('P(theta_hat = x)')
```

* Store the p-value of this test in a variable named
`ans_3_p_value`.


## 4
```{r, echo=F, fig.width=10}
n <- 2

mu_x_0 <- 5
sigma_x <- 2 / sqrt(n)
x_crit <- qnorm(0.95, mu_x_0, sigma_x, lower.tail=F)
mu_x_1 <- x_crit

x <- seq(mu_x_0 - 5*sigma_x, mu_x_1 + 5*sigma_x, 0.01)
fx0 <- dnorm(x, mu_x_0, sigma_x)
fx1 <- dnorm(x, mu_x_1, sigma_x)
d <- data.table(x, fx0, fx1)
d[x <= x_crit, region0 := 'I'] # confidence
d[x > x_crit, region0 := 'II'] # alpha
d[x <= x_crit, region1 := 'III'] # beta
d[x > x_crit, region1 := 'IV'] # power

ggplot(d, aes(x=x)) +
  geom_line(aes(y=fx0)) +
  geom_line(aes(y=fx1)) +
  geom_vline(xintercept=x_crit, linetype=2) +
  geom_ribbon(data=d, aes(x=x, ymin=0, ymax=fx0, fill=region0), alpha=0.25) +
  geom_ribbon(data=d, aes(x=x, ymin=0, ymax=fx0, fill=region0), alpha=0.25) +
  geom_ribbon(data=d, aes(x=x, ymin=0, ymax=fx1, fill=region1), alpha=0.25) +
  geom_ribbon(data=d, aes(x=x, ymin=0, ymax=fx1, fill=region1), alpha=0.25) +
  scale_x_continuous(breaks=c(mu_x_1, mu_x_0), labels=c('H1', 'H0')) +
  ylab('Probability Density') +
  theme(legend.title = element_blank())
```

Please respond by assigning `"confidence"`, `"alpha"`,
`"beta"`, or `"power"` to the variables requested below.
Note that the dashed line in the above figure is the
critical value. Please also note that the colour of region I
is a bit hard to resolve due to the transparency and colour
blurring, so look carefully.

* What quantity does region I in the above plot correspond
to? Store your answer in a variable named `ans_4a`.

* What quantity does region II in the above plot correspond
to? Store your answer in a variable named `ans_4b`.

* What quantity does region III in the above plot correspond
to? Store your answer in a variable named `ans_4c`.

* What quantity does region IV in the above plot correspond
to? Store your answer in a variable named `ans_4d`.

* What is the numeric value (to two decimal places) of the
power in this example? Store your answer in a variable named
`ans_4e`.

* Assuming a type I error rate of 0.05, what is the numeric
value (to two decimal places) of confidence of this example?
Store your answer in a variable named `ans_4f`.

* Assuming that the illustrated distributions correspond to
the distribution of sample means, will increasing the sample
size increase the distance between the mean of H1 and H0
distributions? Store your answer using `ans_4g <- "YES"` or
`ans_4g <- "NO"`.


## Final Check
A common error on these problem sets is to accidentally
erase or overwrite variables from one question with those
from another. Another common error is to name your variables
slightly wrong (e.g., `Ans_1a` instead of `ans_1a`). Yet
another common error is to include a line of code or two
that generates an error, and sometimes this error is serious
enough to prevent most or all of your script from running,
in which case you will lose most or all of your marks for
that assignment. All of these can be a very frustrating way
to lose marks, but you will indeed lose marks if you make
these mistakes, so you need to be very careful!

To ensure that this doesn't happen to you, please run your
entire `.R` script from start to finish. One way to do this
is to use `Code > Run Region > Run All`, but of course there
are shortcuts for everything so do as you wish. If there are
any errors at all when you do this, it is essential that you
address them.

After all of your code has executed from start to finish
without and errors at all, then carefully inspect your
workspace to ensure that the following variables are
defined. A reasonable way to do this is to try to print each
one of these variables to the console. If it is defined,
then it will print without error.

```{r, eval=F, echo=T}
my_name
my_student_id
ans_1_test_stat_obs
ans_1_p_value
ans_2_test_stat_obs
ans_2_p_value
ans_3_p_value
ans_4a
ans_4b
ans_4c
ans_4d
ans_4e
ans_4f
ans_4g
```

Finally, be sure that the file you submit to iLearn is a
`.R` file and nothing else. Any other extension cannot be
marked. This means do not submit `.Rmd`, `.RProj`, `.pdf`,
`.html` or anything else. You must submit a `.R` file.