---
title: "Homework 1"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(data.table)
library(ggplot2)

rm(list=ls())
```

## 0.
* Create a new file named `cogs2020_hw_1.R` and save it in
the `cogs2020` folder that you created during Lecture 1.

* Make the first line of your script `rm(list=ls())`. This
line of code will erase any variable defined before it. Do
not put this line of code after anything important that you
want to keep (e.g., so that I can mark your work).

* After `rm(list=ls())`, create a variable named `my_name`
and set its value equal to a **character** vector (i.e.,
letters surrounded by `""` or by `''`) containing your name.

* After `rm(list=ls())`, create a variable named
`my_student_id` and set its value equal to a **character**
vector (i.e., letters surrounded by `""` or by `''`)
containing your student id.

- So far, here is what your file should look like if your
name is `John Doe` and and your student ID is `12345678`.
```{r, echo=T}
rm(list=ls())
my_name <- "John Doe"
my_student_id <- "12345678"
```

## 1.
a. Define a variable named `ans_1a_x` and set its value to
`10`, a variable named `ans_1a_y` and set its value to `5`,
and a variable named `ans_1a_z` and set its value to the sum
of `ans_1a_x` and `ans_1a_y`. The line of code that you
write to define `ans_1a_z` must include the `+` operator or
you will be marked incorrect.

b. Define a variable named `ans_1b_x` and set its value to
`"10"`, a variable named `ans_1b_y` and set its value to
`"5"`, and a variable named `ans_1b_z` and set its value to
the sum of the *numeric values* of `ans_1b_x` and
`ans_1b_y`. The line of code that you write to define
`ans_1b_z` must include the `+` operator and the
`as.numeric()` function or you will be marked incorrect.

c. Define a variable named `ans_1c` and set its value to the
`logical` result indicating if `ans_1a_x` is equal to
`ans_1a_y`. The line of code that you write to define
`ans_1c` must include the `==` operator or you will be
marked incorrect.

## 2.
a. Create a `vector` named `ans_2a` that contains the
elements `2`, `3`, `5`, `7`, and `11`.

b. Create a `list` named `ans_2b` that contains the elements
`1`, `2`, `"a"`, `"b"` and `TRUE`.

c. Create a `data.frame` named `ans_2c` that contains `v1`,
`v2`, and `v3` as defined below as columns. Make sure the
column names of the `data.frame` you create are equal to the
variable names `v1`, `v2`, and `v3`.
```{r, echo=T}
v1 <- c('I', 'I', 'I', 'I', 'II', 'II', 'II', 'II')
v2 <- c('a', 'a', 'b', 'b', 'c', 'c', 'd', 'd')
v3 <- c(-1.6297880, 
        -1.0738506,  
         0.0299236, 
        -1.5435811, 
        -0.5133278, 
        -1.4716107,
        -1.1986316,
        -1.5548207)
```

## 3.
a. Write a line of code that returns the third element from
the `vector` named `tmp` defined in the following code chunk
and store the result in a variable named `ans_3a`.
```{r, echo=T}
tmp <- c(0.24, 0.015, 1.34, -1.00, -0.15)
```

b. Write a line of code that returns the second element from
the `list` named `tmp` defined in the following code chunk
and store the result in a variable named `ans_3b`. Be sure
that the data type of `ans_3b` is `character` or you will be
marked incorrect.
```{r, echo=T}
tmp <- list(0.5, "W", FALSE)
``` 

c. Write a line of code that returns the column named `v2`
from the `data.frame` named `tmp` defined in the following
code chunk and store the result in a variable named
`ans_3c`. Be sure that the line of code that you write
includes the `$` operator or you will be marked incorrect.
```{r, echo=T}
tmp <- data.frame(v1=c(1, 2, 3),
                  v2=c('A', 'B', 'C'),
                  v3=c(TRUE, TRUE, FALSE))
``` 

## 4.
a. For the following code chunk, what is the data type of
the second element of `tmp`? Store your answer in a variable
named `ans_4a` by copying, pasting, and uncommenting one of
the commented out lines of code below.
```{r, echo=T}
element_1 <- "A"
element_2 <- 2
element_3 <- FALSE
tmp <- c(element_1, element_2, element_3)

# ans_4a <- "character"
# ans_4a <- "numeric"
# ans_4a <- "logical"
``` 

b. For the following code chunk, what is the data type of
the second element of `tmp`? Store your answer in a variable
named `ans_4b` by copying, pasting, and uncommenting one of
the commented out lines of code below.
```{r, echo=T}
element_1 <- "A"
element_2 <- 2
element_3 <- FALSE
tmp <- c(element_1, element_2, element_3)

# ans_4b <- "character"
# ans_4b <- "numeric"
# ans_4b <- "logical"
``` 

## 5.
The function `generate_experiment_result()` defined in the
following code chunk takes `sample_size` as an argument and
returns the results from a silly experiment in which we
sample from a normal distribution.
```{r, echo=T}
set.seed(0)
generate_experiment_result <- function(sample_size) {
  result <- rnorm(sample_size)
  return(result)
}

experiment_result <- generate_experiment_result(10)

experiment_result_mean <- mean(experiment_result)

critical_value <- 1.5
if(experiment_result_mean > critical_value) {
   experiment_decision <- TRUE
} else {
   experiment_decision <- FALSE
}
``` 
a. Is the function `generate_experiment_result ` a
**descriptive statistic**, an **inferential statistic**, a
**random variable**, or a **sample** from a random variable?
Please use one of the following lines of code to save your
answer.
```{r, echo=T}
ans_5a <- "descriptive_statistic"
ans_5a <- "inferential_statistic"
ans_5a <- "random_variable"
ans_5a <- "sample"
``` 

b. Is the variable `experiment_result` a **descriptive
statistic**, an **inferential statistic**, a **random
variable**, or a **sample** from a random variable?
Please use one of the following lines of code to save your
answer.
```{r, echo=T}
ans_5b <- "descriptive_statistic"
ans_5b <- "inferential_statistic"
ans_5b <- "random_variable"
ans_5b <- "sample"
``` 

c. Is the variable `experiment_result_mean` a **descriptive
statistic**, an **inferential statistic**, a **random
variable**, or a **sample** from a random variable?
Please use one of the following lines of code to save your
answer.
```{r, echo=T}
ans_5c <- "descriptive_statistic"
ans_5c <- "inferential_statistic"
ans_5c <- "random_variable"
ans_5c <- "sample"
``` 

d. Is the variable `experiment_decision` a **descriptive
statistic**, an **inferential statistic**, a **random
variable**, or a **sample** from a random variable?
Please use one of the following lines of code to save your
answer.
```{r, echo=T}
ans_5d <- "descriptive_statistic"
ans_5d <- "inferential_statistic"
ans_5d <- "random_variable"
ans_5d <- "sample"
``` 

## 6.
A common error on these problem sets is to accidentally
erase or overwrite variables from one question with those
from another. Another common error is to name your variables
slightly wrong (e.g., `Ans_1a` instead of `ans_1a`). Yet
another common error is to include a line of code or two
that generates an error, and sometimes this error is serious
enough to prevent most or all of your script from running,
in which case you will lose most or all of your marks for
that assignment. All of these can be a very frustrating way
to lose marks, but you will indeed lose marks if you make
these mistakes, so you need to be very careful!

To ensure that this doesn't happen to you, please run your
entire `.R` script from start to finish. One way to do this
is to use `Code > Run Region > Run All`, but of course there
are shortcuts for everything so do as you wish. If there are
any errors at all when you do this, it is essential that you
address them.  

After all of your code has executed from start to finish
without and errors at all, then carefully inspect your
workspace to ensure that the following variables are
defined. A reasonable way to do this is to try to print each
one of these variables to the console. If it is defined,
then it will print without error.
```{r, eval=F, echo=T}
my_name
my_student_id
ans_1a_x
ans_1a_y
ans_1a_z
ans_1b_x
ans_1b_y
ans_1b_z
ans_1c
ans_2a
ans_2b
ans_2c
ans_3a
ans_3b
ans_3c
ans_4a
ans_4b
ans_5a
ans_5b
ans_5c
ans_5d
``` 
