---
title: "Homework 6"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r, echo=F}
library(data.table)
library(ggplot2)
rm(list = ls())
```

**1.** Consider the following data:

```{r, echo=F}
a <- rnorm(5, 10, 2)
b <- rnorm(5, 12, 2)
c <- rnorm(5, 14, 2)
d <- rnorm(5, 10, 2)
d <- data.table(a, b, c, d)
d
```

* Assume that the observations in columns `a` are drawn from
a random variable $A \sim \mathcal{N}(\mu_A, \sigma_A)$.

* Assume that the observations in columns `b` are drawn from
a random variable $B \sim \mathcal{N}(\mu_B, \sigma_B)$.

* Assume that the observations in columns `c` are drawn from
a random variable $C \sim \mathcal{N}(\mu_C, \sigma_C)$.

* Assume that the observations in columns `d` are drawn from
a random variable $D \sim \mathcal{N}(\mu_D, \sigma_D)$.

* Test the hypothesis that $\mu_A$, $\mu_B$, $\mu_C$, and
$\mu_D$ are different. Please populate the following
variables.

```{r, eval=F}
prob_1_theta_hat_x <- theta_hat_x # possible outcomes of your test statistic
prob_1_theta_hat_p <- theta_hat_p # corresponding probabilities of outcomes
prob_1_theta_hat_obs <- theta_hat_obs # observed value of test statistic
prob_1_pval <- p_val
prob_1_theta_hat_crit <- theta_hat_crit # critical value of test statistic
prob_1_decision <- decision
prob_1_r_result <- r_result

# NOTE: when determining the possible outcomes of your test
# statistic, it is okay to truncate. For example, if your
# test statistic has a minimum outcome that is less than 4
# standard deviations away from the mean, then just stop at
# the minimum value.

# NOTE: the population mean of an F distribution (if
# df_denominator > 2) is:

# df_denominator/(df_denominator - 2)
```

**2.** Consider the following sampling distribution of an
$F_{obs}$ test statistic (indicated by the dashed line in
the plot), with $df_{between} = 10$ and $df_{within} = 100$.

```{r, echo=F}
set.seed(1)
x <- seq(0, 4, 0.01)
fx <- df(x, 10, 100)
f_obs <- rf(1, 10, 100)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_line() +
  geom_vline(xintercept = f_obs, linetype=2) +
  scale_x_continuous(breaks=c(0, f_obs, 1, 2, 3, 4))
```

* Are the mean sqaured deviations greater within groups or
are they greater between groups? Please respond with one of
the following:

  * `ans_2a <- "within"` 
  * `ans_2a <- "between"` 
  
* What is the p-value of this test? Please store your answer
in a variable named `ans_2b`.

* What is the critical value of this test? Please store your
answer in a variable named `ans_2c`.


**3.** Consider the motor learning experiment data located
here:

https://crossley.github.io/cogs2020/data/mis/mis_data.csv

* Please consult HW1 for a detailed description of this
data. Please also note that this is a between subjects
design with respect to `condition`. This means that
different subjects are in each level of `condition`. This is
true despite the `subject` numbers being identical. They are
only identical because of an error in how the data were
processed.

* Is the mean `error` per `subject` different across
different levels of `condition`? Answer this question by
performing a one-way ANOVA. Only `prob_3_r_result` will be
marked for this problem.

**4.** Repeat problem 3 but this time use an appropriate
t-test (assuming there is equal variance in different
conditions). For this problem, only `prob_4_r_result` will
be marked for this problem.