---
title: "Homework 4"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 18pt
---
```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
```

```{r setup, include=FALSE, message=F}
knitr::opts_chunk$set(echo = T, collapse=T, warning=F)
```

<style type="text/css">
  body{
  font-size: 18pt;
}

<style type="text/css">
.table {

    width: 40%;
}
</style>

```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)
library(latex2exp)
```

## 0.
* Create a new file named `cogs2020_hw_4.R` and save it in
the `cogs2020` folder that you created during Lecture 1.

* If you have not already installed the `data.table` and the
`ggplot2` packages, do so now. If you use the
`install.packages` function, do not include it in the script
that you hand in.

* Make the first line of your script `library(data.table)`
and the second line `library(ggplot2)`.

* Make the second line of your script `rm(list=ls())`. This
line of code will erase any variable defined before it. Do
not put this line of code after anything important that you
want to keep (e.g., so that I can mark your work).

* After `rm(list=ls())`, create a variable named `my_name`
and set its value equal to a **character** vector (i.e.,
letters surrounded by `""` or by `''`) containing your name.

* After `rm(list=ls())`, create a variable named
`my_student_id` and set its value equal to a **character**
vector (i.e., letters surrounded by `""` or by `''`)
containing your student id.

* So far, here is what your file should look like if your
name is `John Doe` and and your student ID is `12345678`.
Note that the following code chunk also includes a few lines
to create a `data.table` named `d` that many of the problems
below require you manipulate.
```{r, echo=T}
# load the packages we will need
library(data.table)
library(ggplot2)

# clean session
rm(list=ls())

# basic id info
my_name <- "John Doe"
my_student_id <- "12345678"

# Please include the following line of code as well
set.seed(0)
```


## 1.

Consider the following two random variables:

* $X\sim\mathcal{N}(\mu=0,\sigma=1)$ 

* $Y{\sim}Binom(n=10,p=0.2)$ 

The $X$ probability distribution (pdf since $X$ is
continuous) and the $Y$ probability distribution (pmf since
$Y$ is discrete) are illustrated below. The probability
corresponding to $x{\leq}1$ and $y{\leq}3$
(`lower.tail=TRUE`) are coloured blue and the the
probability corresponding to $x{>}1$ and $y{>}3$
(`lower.tail=FALSE`) are coloured green. This figure is only
here to help you think about what is being asked below.

```{r, echo=F, fig.width=10}
x <- seq(-4, 4, 0.01)
fx <- dnorm(x)
d <- data.table(x, fx)
ga <- ggplot(d, aes(x, fx)) +
  geom_line() +
  geom_segment(aes(x=1, xend=1, y=0, yend=dnorm(1)), linetype=2) +
  geom_ribbon(data=d[x<=1],
              aes(x=x, ymin=0, ymax=fx),
              fill='blue',
              alpha=0.25) +
  geom_ribbon(data=d[x>1],
              aes(x=x, ymin=0, ymax=fx),
              fill='green',
              alpha=0.25) +
  annotate("text", x=-0.5, y=0.05, label=TeX("$P(X\\leq x)$")) +
  annotate("text", x=1.5, y=0.05, label=TeX("$P(X>x)$"))+
  ylab('f(x)')

n <- 10
p <- 0.2
y <- seq(0, n, 1)
fy <- dbinom(y, n, p)
d <- data.table(y, fy)
gb <- ggplot(d, aes(y, fy)) +
  geom_segment(data=d[y<=3], aes(x=y, xend=y, y=0, yend=dbinom(y,n,p)), colour='blue') +
  geom_segment(data=d[y>3], aes(x=y, xend=y, y=0, yend=dbinom(y,n,p)), colour='green') +
  annotate("text", x=3, y=0.21, label=TeX("$P(Y\\leq y)$")) +
  annotate("text", x=4, y=0.1, label="P(Y>y)") +
  scale_x_continuous(breaks=0:10) +
  ylab('f(y)')

ggarrange(ga, gb, ncol=2)
```

Are the following statements `TRUE` or `FALSE`?

**(a)** $p(X{<}x)=p(X{\leq}x)$ for all $x$.
```{r}
# uncomment the correct answer
# ans_1a <- TRUE
# ans_1a <- FALSE
```

**(b)** $p(Y < y) = p(Y\leq y)$ for all $y$.
```{r}
# uncomment the correct answer
# ans_1b <- TRUE
# ans_1b <- FALSE
```

**(c)** $p(X < x) = 1 - p(X > x)$ for all $x$.
```{r}
# uncomment the correct answer
# ans_1c <- TRUE
# ans_1c <- FALSE
```

**(d)** $p(Y < y) = 1 - p(Y > y)$ for all $y$.
```{r}
# uncomment the correct answer
# ans_1d <- TRUE
# ans_1d <- FALSE
```

## 2.

Consider the following random variables:

* $X{\sim}\mathcal{N}(\mu_X=15,\sigma_X=2.74)$ 

* $Y{\sim}Binom(n=30,p=0.5)$.

The $X$ and $Y$ distributions are illustrated in the
following figure:

```{r, echo=F, fig.width=10}
x <- seq(0, 30, 0.01)
fx <- dnorm(x, 15, 2.74)
d <- data.table(x, fx)
gx <- ggplot(d, aes(x, fx)) +
  geom_line() +
  ylab('f(x)')

n <- 30
p <- 0.5
y <- seq(0, 30, 1)
fy <- dbinom(y, n, p)
d <- data.table(y, fy)
gy <- ggplot(d, aes(y, fy)) +
  geom_segment(aes(x=y, xend=y, y=0, yend=fy)) + 
  ylab('f(y)')

ggarrange(gx, gy, ncol=2)
```

For all of the following use built-in `R` functions to
compute the requested quantities. The functions are covered
in the lecture notes this week.

**(a)** Compute $P(X > 17)$ and store it in a variable
named `ans_2a`.

**(b)** Compute $P(X \leq 14)$ and store it in a variable
named `ans_2b`.

**(c)** Compute $P(X \geq 20)$ and store it in a variable
named `ans_2c`.

**(d)** Compute $P(X < 13)$ and store it in a variable
named `ans_2d`.

**(e)** Compute $P(Y > 17)$ and store it in a variable
named `ans_2e`.

**(f)** Compute $P(Y \leq 14)$ and store it in a variable
named `ans_2f`.

**(g)** Compute $P(Y \geq 20)$ and store it in a variable
named `ans_2g`.

**(h)** Compute $P(Y < 13)$ and store it in a variable
named `ans_2h`.

**(i)** How do all of the above quantities computed with
respect to $X$ compare to those computed with respect to
$Y$? No points for this one. Just think about it.

<!-- **(e)** Compute $P(12.22 < X < 24.55)$ and store it in a -->
<!-- variable named `ans_2e`. -->

<!-- **(f)** Compute $x$ such that $P(X>x) = .9$ and store it in -->
<!-- a variable named `ans_2f`. -->

<!-- **(g)** Compute $x$ such that $P(X \leq x) = .8$ and store -->
<!-- it in a variable named `ans_2g`. -->


## 3.

Consider the following plots:

```{r, echo=F, fig.height=14, fig.width=10}
xcritvec <- c(17, 14, 20, 13)

mu <- 15
sigma <- 2.74

x <- seq((mu-sigma*4), (mu+sigma*4), 0.01)
fxd <- dnorm(x, mu, sigma)

fxp <- pnorm(x, mu, sigma, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[1]
ga1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x>xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)')

ga2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)')

fxp <- pnorm(x, mu, sigma, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[2]
gb1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x <= xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)')

gb2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=T),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=T),
               yend=pnorm(xcrit, mu, sigma, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)')

fxp <- pnorm(x, mu, sigma, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[3]
gc1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x >= xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)')

gc2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)')

fxp <- pnorm(x, mu, sigma, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[4]
gd1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x < xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)')

gd2n <- ggplot(d, aes(x, 1-fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab(TeX('P(X\\leq x)'))

n <- 30
p <- 0.5
x <- seq(0, n, 1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[1]
ga1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x>xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x<=xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))

ga2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit, 
               xend=xcrit, 
               y=0, 
               yend=pbinom(xcrit, n, p, lower.tail=F), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pbinom(xcrit, n, p, lower.tail=F),
               yend=pbinom(xcrit, n, p, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  scale_x_continuous(breaks=seq(0,n,2))

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[2]
gb1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x<=xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x>xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))

gb2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit, 
               xend=xcrit, 
               y=0, 
               yend=pbinom(xcrit, n, p, lower.tail=T), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pbinom(xcrit, n, p, lower.tail=T),
               yend=pbinom(xcrit, n, p, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  scale_x_continuous(breaks=seq(0,n,2))

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[3]
gc1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x>=xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x<xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))

gc2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit-1, 
               xend=xcrit-1, 
               y=0, 
               yend=pbinom(xcrit-1, n, p, lower.tail=F), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit-1,
               y=pbinom(xcrit-1, n, p, lower.tail=F),
               yend=pbinom(xcrit-1, n, p, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  scale_x_continuous(breaks=seq(0,n,2))

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[4]
gd1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x<xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x>=xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))

gd2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit-1, 
               xend=xcrit-1, 
               y=0, 
               yend=pbinom(xcrit-1, n, p, lower.tail=T), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit-1,
               y=pbinom(xcrit-1, n, p, lower.tail=T),
               yend=pbinom(xcrit-1, n, p, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  scale_x_continuous(breaks=seq(0,n,2))

list_1 <- list(ga1n, gb1n, gc1n, gd1n, ga1b, gb1b, gc1b, gd1b)
list_2 <- list(ga2n, gb2n, gc2n, gd2n, ga2b, gb2b, gc2b, gd2b)
plot_list_list <- list(list_1, list_2)

set.seed(2)
inds1 <- sample(1:8, 8)
inds2 <- sample(rep(1:2, 4), 8)

plot_list <- list()
for(i in 1:8) {
  plot_list[[i]] <- plot_list_list[[inds2[i]]][[inds1[i]]]
}

plot_labels <- list('I', 'II', 'III', 'IV',
                    'V', 'VI', 'VII', 'VIII')

ggarrange(
  plotlist=plot_list,
  labels=plot_labels,
  nrow=4,
  ncol=2)
```

For each of the following problems, please respond by
setting the appropriate variable to one of `"I"`, `"II"`,
`"III"`, `"IV"`, `"V"`, `"VI"`, `"VII"`, or `"VIII"`.
When responding to the questions below, please note that 
the x and y axis labels in the above plots all use 'X' or 
'x' but may correspond to either the X or the Y random variable 
from problem 2.

**(a)** Which plot corresponds to the quantity requested 
in problem 2a? Please store your answer in a variable named
`ans_3a`.

**(b)** Which plot corresponds to the quantity requested 
in problem 2b? Please store your answer in a variable named
`ans_3b`.

**(c)** Which plot corresponds to the quantity requested 
in problem 2c? Please store your answer in a variable named
`ans_3c`.

**(d)** Which plot corresponds to the quantity requested 
in problem 2d? Please store your answer in a variable named
`ans_3d`.

**(e)** Which plot corresponds to the quantity requested 
in problem 2e? Please store your answer in a variable named
`ans_3e`.

**(f)** Which plot corresponds to the quantity requested 
in problem 2f? Please store your answer in a variable named
`ans_3f`.

**(g)** Which plot corresponds to the quantity requested 
in problem 2g? Please store your answer in a variable named
`ans_3g`.

**(h)** Which plot corresponds to the quantity requested 
in problem 2h? Please store your answer in a variable named
`ans_3h`.


## 4.
Consider the random variable $X{\sim}F(d_1,d_2)$ with fixed
parameters $d_1=5$, $d_2=100$.  Given these values of $d_1$
and $d_2$, the following holds true:

$$
\text{if } X \sim F(d_1,d_2)\text{ then:}\\
\mathbb{E}[X] = \frac{d_2}{d_2 - 2} \\ 
\mathbb{V}\text{ar}[X] = \frac{2 d_2^2 (d_1 + d_2 - 2)}{d_1 (d_2 - 2)^2 (d_2 - 4)}
$$

You might need numerical values for $\mathbb{E}[X]$ and
$\mathbb{V}\text{ar}[X]$ later in this problem. If you do,
you can compute them with the following code chunk:

```{r}
d1 <- 5
d2 <- 100

mu_x <- (d2/(d2-2))
sig_x <- (2*(d2^2)*(d1+d2-2))/(d1*(d2-2)^2*(d2-4))
```

We next perform 6 experiments. In each experiment, we draw
1000 samples from $X$ and plot a histogram for the results
of each experiment. The results of these experiments are
shown in the following histograms.

```{r, echo=F, include=T}
d1 <- 5
d2 <- 100

# Define the number of experiments to run
n_exp <- 6

# Define the number of samples to draw per experiment
n <- 1000

# Create an empty list. We will later use this list to store
# the results of each experiment
storage_list <- list()

# Run through a for loop and in each iteration of the for
# loop, draw n samples from the random variable X, and store
# the result in the storage list defined just above.
for(i in 1:n_exp) {
  
  # We will use the function `rf()` to draw random samples
  # from X. However, I need you all to work with the same
  # random samples as me (for grading purposes), so we use
  # the `set.seed()` function to ensure that we all get the
  # same "random" numbers.
  set.seed(i)
  
  # Generate a random sample using the `rf()` function
  random_sample <- rf(n, d1, d2)
  
  # we will want things in data.table format later, so might
  # as well get started on it now.
  result <- data.table(exp=i, x=random_sample)
  
  # Store the result in our storage list
  storage_list[[i]] <- result
}

# Storage_list contains a list of data.tables. Each
# data.table in this list has one column named exp (this is
# an indicator column that indicates the experiment from
# which data was obtained) and another named x (this column
# contains the data sampled in each experiment). We next use
# the data.table function `rbindlist()` to combine this list
# of data.tables into a single data.table.
d <- rbindlist(storage_list)

# We now plot histograms to show the results of each
# experiment.
ggplot(d, aes(x)) +
  geom_histogram(bins=50) +
  facet_wrap(~exp)
```

**(a)**Do the histograms appear Normally distributed? If
not, what is the main difference? Please respond with one of
the following:

* `ans_4a <- "Skewed"`
* `ans_4a <- "Normal"`

We next estimate the distribution of sample means as a
function of the number of samples `n` that were drawn per
experiment. We will perform 6 experiments in which we draw
2, 5, 10, 15, 30, and 50 samples from $X$ (each of these
numbers corresponds to one experiment). Each of these
experiments, if performed once, will give us one sample
mean. To examine the distribution of sample means, we repeat
each experiment 1000 times. This will give us 1000 sample
means per experiment. For each experiment, we estimate the
distribution of sample means with a histogram and make a
single figure with separate panels for each experiment to
show our results.

```{r, echo=F, include=T}
n_exp <- 1000
n <- c(2, 5, 10, 15, 30, 50)

storage_list <- list()
for(j in 1:length(n)) {
  for(i in 1:n_exp) {
    set.seed(i + j*n_exp)
    random_sample <- rf(n[j], d1, d2)
    sample_mean <- mean(random_sample)
    result <- data.table(exp=i, n=n[j], x=sample_mean)
    storage_list[[i + j*n_exp]] <- result
  }
}

d <- rbindlist(storage_list)

ggplot(d, aes(x)) +
  geom_histogram(bins=50) +
  facet_wrap(~n)
```

**(b)** What happens to the variance of the distribution of
sample means as $n$ increases?

* `ans_4b <- "Stays the same"`
* `ans_4b <- "Increases"`
* `ans_4b <- "Decreases"`

**(c)** What happens to the skew of the distribution of
sample means as $n$ increases?

* `ans_4c <- "Stays the same"`
* `ans_4c <- "Increases"`
* `ans_4c <- "Decreases"`

**(d)** If $X{\sim}F(d_1,d_2)$ as defined at the top of the
problem, and $n=1000$, what is the population mean of
$\bar{X}$ (the distribution of sample means)? Store your
answer in a variable named `ans_4d`.

**(e)** If $X{\sim}F(d_1,d_2)$ as defined at the top of the
problem, and $n=1000$, what is the population variance of
$\bar{X}$ (the distribution of sample means)? Store your
answer in a variable named `ans_4e`.


## 5. Final Check
A common error on these problem sets is to accidentally
erase or overwrite variables from one question with those
from another. Another common error is to name your variables
slightly wrong (e.g., `Ans_1a` instead of `ans_1a`). Yet
another common error is to include a line of code or two
that generates an error, and sometimes this error is serious
enough to prevent most or all of your script from running,
in which case you will lose most or all of your marks for
that assignment. All of these can be a very frustrating way
to lose marks, but you will indeed lose marks if you make
these mistakes, so you need to be very careful!

To ensure that this doesn't happen to you, please run your
entire `.R` script from start to finish. One way to do this
is to use `Code > Run Region > Run All`, but of course there
are shortcuts for everything so do as you wish. If there are
any errors at all when you do this, it is essential that you
address them.

After all of your code has executed from start to finish
without and errors at all, then carefully inspect your
workspace to ensure that the following variables are
defined. A reasonable way to do this is to try to print each
one of these variables to the console. If it is defined,
then it will print without error.

```{r, eval=F, echo=T}
my_name
my_student_id
ans_1a
ans_1b
ans_1c
ans_1d
ans_2a
ans_2b
ans_2c
ans_2d
ans_2e
ans_2f
ans_2g
ans_2h
ans_3a
ans_3b
ans_3c
ans_3d
ans_3e
ans_3f
ans_3g
ans_3h
ans_4a
ans_4b
ans_4c
ans_4d
ans_4e
```

Finally, be sure that the file you submit to iLearn is a
`.R` file and nothing else. Any other extension cannot be
marked. This means do not submit `.Rmd`, `.RProj`, `.pdf`,
`.html` or anything else. You must submit a `.R` file.