---
title: "Homework 1"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(data.table)
library(ggplot2)

rm(list=ls())
```

## 0.
* Create a new file named `cogs2020_hw_1.R` and save it in
the `cogs2020` folder that you created during Lecture 1.

* Make the first line of your script `rm(list=ls())`. This
line of code will erase any variable defined before it. Do
not put this line of code after anything important that you
want to keep (e.g., so that I can mark your work).

* After `rm(list=ls())`, create a variable named `my_name`
and set its value equal to a **character** vector (i.e.,
letters surrounded by `""` or by `''`) containing your name.

* After `rm(list=ls())`, create a variable named
`my_student_id` and set its value equal to a **character**
vector (i.e., letters surrounded by `""` or by `''`)
containing your student id.

- So far, here is what your file should look like if your
name is `John Doe` and and your student ID is `12345678`.
```{r, echo=T}
rm(list=ls())
my_name <- "John Doe"
my_student_id <- "12345678"
```

## 1.
a. Define a variable named `ans_1a_x` and set its value to
`10`, a variable named `ans_1a_y` and set its value to `5`,
and a variable named `ans_1a_z` and set its value to the sum
of `ans_1a_x` and `ans_1a_y`. The line of code that you
write to define `ans_1a_z` must include the `+` operator or
you will be marked incorrect.

b. Define a variable named `ans_1b_xx` and set its value to
`"10"`, a variable named `ans_1b_yy` and set its value to
`"5"`, and a variable named `zz` and set its value to the
sum of the *numeric values* of `ans_1b_xx` and `ans_1b_yy`.
The line of code that you write to define `zz` must include
the `+` operator and the `as.numeric()` function or you will
be marked incorrect.

c. Define a variable named `x_equal_y` and set its value to
the `logical` result indicating if `x` is equal to `y`. The
line of code that you write to define `x_equal_y` must
include the `==` operator or you will be marked incorrect.

## 2.
a. Create a `vector` named `ans_2a` that contains the
elements `2`, `3`, `5`, `7`, and `11`.

b. Create a `list` named `ans_2b` that contains the elements
`1`, `2`, `"a"`, `"b"` and `TRUE`.

c. Create a `data.frame` named `ans_2c` that contains `v1`,
`v2`, and `v3` as defined below as columns. Make sure the
column names of the `data.frame` you create are equal to the
variable names `v1`, `v2`, and `v3`.
```{r, echo=T}
v1 <- c('I', 'I', 'I', 'I', 'II', 'II', 'II', 'II')
v2 <- c('a', 'a', 'b', 'b', 'c', 'c', 'd', 'd')
v3 <- c(-1.6297880, 
        -1.0738506,  
         0.0299236, 
        -1.5435811, 
        -0.5133278, 
        -1.4716107,
        -1.1986316,
        -1.5548207)
```

## 3.
a. Write a line of code that returns the third element from
the `vector` named `tmp` defined in the following code chunk
and store the result in a variable named `ans_3a`.
```{r, echo=T}
tmp <- c(0.24, 0.015, 1.34, -1.00, -0.15)
```

b. Write a line of code that returns the second element from
the `list` named `tmp` defined in the following code chunk
and store the result in a variable named `ans_3b`. Be sure
that the data type of `ans_3b` is `character` or you will be
marked incorrect.
```{r, echo=T}
tmp <- list(0.5, "W", FALSE)
``` 

c. Write a line of code that returns the column named `v2`
from the `data.frame` named `tmp` defined in the following
code chunk and store the result in a variable named
`ans_3c`. Be sure that the line of code that you write
includes the `$` operator or you will be marked incorrect.
```{r, echo=T}
tmp <- data.frame(v1=c(1, 2, 3),
                  v2=c('A', 'B', 'C'),
                  v3=c(TRUE, TRUE, FALSE))
``` 