---
title: "Homework 3"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 18pt
---
```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
```

<style type="text/css">
  body{
  font-size: 18pt;
}

<style type="text/css">
.table {

    width: 40%;
}
</style>

```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)
library(latex2exp)
```

## 0.
* Create a new file named `cogs2020_hw_3.R` and save it in
the `cogs2020` folder that you created during Lecture 1.

* If you have not already installed the `data.table` and the
`ggplot2` packages, do so now. If you use the
`install.packages` function, do not include it in the script
that you hand in.

* Make the first line of your script `library(data.table)`
and the second line `library(ggplot2)`.

* Make the second line of your script `rm(list=ls())`. This
line of code will erase any variable defined before it. Do
not put this line of code after anything important that you
want to keep (e.g., so that I can mark your work).

* After `rm(list=ls())`, create a variable named `my_name`
and set its value equal to a **character** vector (i.e.,
letters surrounded by `""` or by `''`) containing your name.

* After `rm(list=ls())`, create a variable named
`my_student_id` and set its value equal to a **character**
vector (i.e., letters surrounded by `""` or by `''`)
containing your student id.

* So far, here is what your file should look like if your
name is `John Doe` and and your student ID is `12345678`.
Note that the following code chunk also includes a few lines
to create a `data.table` named `d` that many of the problems
below require you manipulate.
```{r, echo=T}
# load the packages we will need
library(data.table)
library(ggplot2)

# clean session
rm(list=ls())

# basic id info
my_name <- "John Doe"
my_student_id <- "12345678"

# Please include the following line of code as well
set.seed(0)
```

## 1.
Consider an experiment in which two four-sided die are
thrown.

**(a)** Create a `data.table` named `ans_1a` with two
columns labelled `die1` and `die2` where these columns
encode the **sample space** of this experiment. See below
for **only the first** eight rows of the resulting
`data.table`.

```{r echo=FALSE}
die1 <- rep(c(1, 2, 3, 4), 4)
die2 <- rep(c(1, 2, 3, 4), each=4)
ans_1a <- data.table(die1, die2)
ans_1a[1:8]
```

**(b)** Create a deep copy of `ans_1a` called `ans_1b` and
then add a column to `ans_1b` called `X` and set it equal to
the sum of the corresponding `die1` and `die2` values on a
per row basis.

**(c)** Let $X$ be the random variable defined by the `X`
column in `ans_1b`. What is $P(X<5)$? Store your result in a
variable named `ans_1c`.

**(d)** Is the event defined as the set of all outcomes for
which $X=2$ an elementary event? Store your answer (`"YES"`
or `"NO"`) in a variable named `ans_1d`.


## 2.
For the following probability distribution:

<div width:50px>
| $x$    |  2 |  3 |  4 |  5 |  6 |
|:-------|:---|:---|:---|:---|:---|
| $f(x)$ | .1 | .3 | .3 | .2 | .1 |
</div>

**(a)** Is $f(x)$ a valid probability distribution? Store
your answer (`"YES"` or `"NO"`) in a variable named
`ans_2a`.

**(b)** Calculate $E(X)$ store the result in a variable
named `ans_2b`.

**(c)** Calculate $P(X \geq 4)$ store the result in a
variable named `ans_2c`.

**(d)** Calculate $P(2 < X \leq 4)$ and store the result in
a variable named `ans_2d`.


## 3.
Given the following probability distribution:

```{r, echo=FALSE}
x <- c(0, 1, 2, 3)
fx <- c(0.3, 0.4, 0.2, 0.1)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_point() +
  geom_segment(aes(x=x, y=0, xend=x, yend=fx)) +
  xlab("x") +
  ylab("P(X=x)")
```

Find:

**(a)** Calculate $P(X \geq 2)$ and store your result in a
variable named `ans_3a`.

**(b)** Calculate $P(0 < X \leq 2)$ and store your result in
a variable named `ans_3b`.

**(c)** Calculate $E(X)$ and store your result in a variable
named `ans_3c`.

**(d)** $Var(X)$ and store your result in a variable named
`ans_3d`.

**(e)** $sd(X)$ and store your result in a variable named
`ans_3e`.

## 4.
Consider the probability distributions for random vairbale
$X$ and random variable $Y$ defined below.
```{r, echo=F, fig.width=10}
x <- seq(-4, 4, 0.01)
fx <- dnorm(x)
d <- data.table(x, fx)
ga <- ggplot(d, aes(x, fx)) +
  geom_line() +
  ylab('f(x)')

n <- 2
p <- 0.2
y <- seq(0, n, 1)
fy <- dbinom(y, n, p)
d <- data.table(y, fy)
gb <- ggplot(d, aes(y, fy)) +
  geom_segment(data=d, aes(x=y, xend=y, y=0, yend=dbinom(y,n,p))) +
  scale_x_continuous(breaks=-5:5) +
  scale_y_continuous(breaks=fy) +
  ylab('f(y)')

ggarrange(ga, gb, ncol=2)
```

**a.** What is the population mean of $X$? Store your answer
in a variable named `ans_4a`.

**b.** What is the population mean of $Y$? Sore your answer
in a variable named `ans_4b`.

```{r, echo=F, fig.width=10}
x1 <- rnorm(100)
x2 <- rnorm(100, 10)
y1 <- rbinom(100, n, 0.2)
y2 <- rbinom(100, n, 0.5)
d <- data.table(x1, x2, y1, y2)
d <- melt(d)

ggplot(data=d, aes(value)) +
  geom_histogram(binwidth=0.1) +
  facet_wrap(~variable, scales='free') +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )
```

**c.** Which of the histograms above most likely corresponds
to a sample from $X$? Please store your answer as one of the
following:

* `ans_4c <- 'upper left'`
* `ans_4c <- 'lower left'`
* `ans_4c <- 'upper right'`
* `ans_4c <- 'lower right'`

**d.** Which of the histograms above most likely corresponds
to a sample from $X$? Please store your answer as one of the
following:

* `ans_4d <- 'upper left'`
* `ans_4d <- 'lower left'`
* `ans_4d <- 'upper right'`
* `ans_4d <- 'lower right'`

## 5.
**a.** Use the `fread` function in the `data.table` library
to read
`https://crossley.github.io/cogs2020/data/criterion_learning/crit_learn.csv`
into a `data.table` and store the result in a variable named
`ans_5a`.

**b.** Create a new `data.table` that contains only rows of
`ans_5a` for which `cnd` is `%in%` `c('Delay', 'Long ITI',
'Short ITI')`, and only the mean of the `t2c` column grouped
by `sub` and `cnd`. Store your result in a variable named
`ans_5b`.

**c.** Is `ans_5b` in wide or long format? Please use one of
the following lines to store your result in a variable named
`ans_4c`:

* `ans_5c <- 'wide'`
* `ans_5c <- 'long'`

**d.** If `ans_5b` is in long format, then convert it to
wide format. If it is already wide format, then convert it
to long format. Store your result in a variable named
`ans_5d`. Please also be careful to not alter `ans_5b` in
the process.

**e.** If `ans_5d` is in long format, then convert it to
wide format. If it is already wide format, then convert it
to long format. Store your result in a variable named
`ans_5e`. Please also be careful to not alter `ans_5d` in
the process.

**f.** Please recreate the following figure using `ans_5b`
as the source of data being plotted. Store the result of
your call to the `ggplot()` function in a variable named
`ans_5f`.

```{r, echo=F, include=T}
d <- fread('https://crossley.github.io/cogs2020/data/criterion_learning/crit_learn.csv')
d <- d[cnd %in% c('Delay', 'Long ITI', 'Short ITI'), .(t2c_mean=mean(t2c)), .(sub, cnd)]
ggplot(data=d, aes(x=cnd, y=t2c_mean)) +
  geom_violin() +
  geom_point()
```

## 6. Final Check
A common error on these problem sets is to accidentally
erase or overwrite variables from one question with those
from another. Another common error is to name your variables
slightly wrong (e.g., `Ans_1a` instead of `ans_1a`). Yet
another common error is to include a line of code or two
that generates an error, and sometimes this error is serious
enough to prevent most or all of your script from running,
in which case you will lose most or all of your marks for
that assignment. All of these can be a very frustrating way
to lose marks, but you will indeed lose marks if you make
these mistakes, so you need to be very careful!

To ensure that this doesn't happen to you, please run your
entire `.R` script from start to finish. One way to do this
is to use `Code > Run Region > Run All`, but of course there
are shortcuts for everything so do as you wish. If there are
any errors at all when you do this, it is essential that you
address them.

After all of your code has executed from start to finish
without and errors at all, then carefully inspect your
workspace to ensure that the following variables are
defined. A reasonable way to do this is to try to print each
one of these variables to the console. If it is defined,
then it will print without error.

```{r, eval=F, echo=T}
my_name
my_student_id
ans_1a
ans_1b
ans_1c
ans_1d
ans_2a
ans_2b
ans_2c
ans_2d
ans_3a
ans_3b
ans_3c
ans_3d
ans_3e
ans_4a
ans_4b
ans_4c
ans_4d
ans_5a
ans_5b
ans_5c
ans_5d
ans_5e
```
