---
title: "Homework 3"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

<div style="color:#990000">
**Note that this homework is not yet finalised. I am posting
it here for practice purposes only.**
<br></br>
</div>

<style type="text/css">
.table {

    width: 40%;

***
</style>


```{r, echo=F}
library(data.table)
library(ggplot2)
library(ggpubr)
```


## Problem 1

$p(X<x)=p(X≤x)$ for continuous distributions but not for
discrete distributions.

**(a)** Verify that $p(X<x)=p(X \leq x)$ for:

  * $X\sim\mathcal{N}(\mu,\sigma)$ 
  * $x=1$ 
  * $\mu=0$ 
  * $\sigma=1$ 
  
  * Do this by performing the following steps:

  * Define a variable `p1` and set its value to $p(X<x)$.
  Use `lower.tail=FALSE` to do so.
  
  * Define a variable `p2` and set its value to
  $p(X{\leq}x$. Use `lower.tail=TRUE` to do so.
  
  * Define a variable `ans_1a` and set its value to `TRUE`
  if `p1` is equal to `p2` and set its value to `FALSE`
  otherwise.
  
  * The $X$ probability distribution (pdf since X is
  continuous) and the probabilities given by
  `lower.tail=TRUE` and `lower.tail=FALSE` are illustrated
  in the following figure. This figure is only here to help
  you think about what is being asked.

```{r, echo=F}
x <- seq(-5, 5, 0.01)
fx <- dnorm(x)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_line() +
  geom_segment(aes(x=1, xend=1, y=0, yend=dnorm(1)), linetype=2) +
  geom_ribbon(data=d[x<=1],
              aes(x=x, ymin=0, ymax=fx),
              fill='blue',
              alpha=0.25) +
  geom_ribbon(data=d[x>1],
              aes(x=x, ymin=0, ymax=fx),
              fill='green',
              alpha=0.25) +
  annotate("text", x=-0.5, y=0.05, label="P(X<=x)") +
  annotate("text", x=1.5, y=0.05, label="P(X>x)") +
  ylab('f(x)')
```


**(b)** Verify that $p(X<x){\neq}p(X{\leq}x)$ for:

  * $X{\sim}binom(n,p)$ 
  
  * $x=3$
  
  * $n=10$
  
  * $p=0.2$. 

  * Do this by performing the following steps:

  * Define a variable `p1` and set its value to $p(X<x)$.
  Use `lower.tail=FALSE` to do so.
  
  * Define a variable `p2` and set its value to
  $p(X{\leq}x$. Use `lower.tail=TRUE` to do so.
  
  * Define a variable `ans_1b` and set its value to `TRUE`
  if `p1` is equal to `p2` and set its value to `FALSE`
  otherwise.
  
  * The $X$ probability distribution (pdf since X is
  continuous) and the probabilities given by
  `lower.tail=TRUE` and `lower.tail=FALSE` are illustrated
  in the following figure. This figure is only here to help
  you think about what is being asked.

```{r, echo=F}
n <- 10
p <- 0.2
x <- seq(0, n, 1)
fx <- dbinom(x, n, p)
d <- data.table(x, fx)
ggplot(d, aes(x, fx)) +
  geom_segment(data=d[x<=3], aes(x=x, xend=x, y=0, yend=dbinom(x,n,p)), colour='blue') +
  geom_segment(data=d[x>3], aes(x=x, xend=x, y=0, yend=dbinom(x,n,p)), colour='green') +
  annotate("text", x=3, y=0.21, label="P(X<=x)") +
  annotate("text", x=4, y=0.1, label="P(X>x)") +
  scale_x_continuous(breaks=0:10) +
  ylab('f(x)')
```

## Problem 2

Consider the random variables
$X{\sim}\mathcal{N}(\mu_X=15,\sigma_X=2.74$ and
$Y{\sim}Binomial(n=30,p=0.5)$.

$X$ and $Y$ are described as illustrated in the following
figure:

```{r, echo=F}
x <- seq(0, 30, 0.01)
fx <- dnorm(x, 15, 2.74)
d <- data.table(x, fx)
gx <- ggplot(d, aes(x, fx)) +
  geom_line() +
  ylab('f(x)')

n <- 30
p <- 0.5
y <- seq(0, 30, 1)
fy <- dbinom(y, n, p)
d <- data.table(y, fy)
gy <- ggplot(d, aes(y, fy)) +
  geom_segment(aes(x=y, xend=y, y=0, yend=fy)) + 
  ylab('f(y)')

ggarrange(gx, gy, ncol=2)
```

**(a)** Compute $P(X > 17)$ and store it in a variable
named `ans_2a`.

**(b)** Compute $P(X \leq 14)$ and store it in a variable
named `ans_2b`.

**(c)** Compute $P(X \geq 20)$ and store it in a variable
named `ans_2c`.

**(d)** Compute $P(X < 13)$ and store it in a variable
named `ans_2d`.

**(e)** Compute $P(Y > 17)$ and store it in a variable
named `ans_2a`.

**(f)** Compute $P(Y \leq 14)$ and store it in a variable
named `ans_2b`.

**(g)** Compute $P(Y \geq 20)$ and store it in a variable
named `ans_2c`.

**(h)** Compute $P(Y < 13)$ and store it in a variable
named `ans_2d`.

**(i)** How do all of the above quantities computed with
respect to $X$ compare to those computed with respect to
$y$? No points for this one. Just think about it.

<!-- **(e)** Compute $P(12.22 < X < 24.55)$ and store it in a -->
<!-- variable named `ans_2e`. -->

<!-- **(f)** Compute $x$ such that $P(X>x) = .9$ and store it in -->
<!-- a variable named `ans_2f`. -->

<!-- **(g)** Compute $x$ such that $P(X \leq x) = .8$ and store -->
<!-- it in a variable named `ans_2g`. -->

## Problem 3

**(a)** Which of the following plots corresponds to the
quantity requested in problem 2a? Please respond by creating
a variable named `ans_3a` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(b)** Which of the following plots corresponds to the
quantity requested in problem 2b? Please respond by creating
a variable named `ans_3b` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(c)** Which of the following plots corresponds to the
quantity requested in problem 2c? Please respond by creating
a variable named `ans_3c` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2d? Please respond by creating
a variable named `ans_3b` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2e? Please respond by creating
a variable named `ans_3e` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2f? Please respond by creating
a variable named `ans_3f` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2g? Please respond by creating
a variable named `ans_3g` and setting its value to one of
`I`, `II`, `III`, or `IV`.

**(d)** Which of the following plots corresponds to the
quantity requested in problem 2h? Please respond by creating
a variable named `ans_3h` and setting its value to one of
`I`, `II`, `III`, `IV`, `V`, `VI`, `VII`, or `VIII`.

```{r, echo=F}
xcritvec <- c(17, 14, 20, 13)

mu <- 15
sigma <- 2.74

x <- seq((mu-sigma*4), (mu+sigma*4), 0.01)
fxd <- dnorm(x, mu, sigma)

fxp <- pnorm(x, mu, sigma, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[1]
ga1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x>xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

ga2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  theme(aspect.ratio=1)

fxp <- pnorm(x, mu, sigma, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[2]
gb1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x <= xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

gb2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=T),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=T),
               yend=pnorm(xcrit, mu, sigma, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  theme(aspect.ratio=1)

fxp <- pnorm(x, mu, sigma, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[3]
gc1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x >= xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

gc2n <- ggplot(d, aes(x, fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  theme(aspect.ratio=1)

fxp <- pnorm(x, mu, sigma, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[4]
gd1n <- ggplot(d, aes(x, fxd)) +
  geom_line() +
  geom_ribbon(data=d[x < xcrit],
              aes(x=x, ymin=0, ymax=fxd),
              fill='blue',
              alpha=0.25) +
  ylab('f(x)') +
  theme(aspect.ratio=1)

gd2n <- ggplot(d, aes(x, 1-fxp)) +
  geom_line() +
  geom_segment(x=xcrit,
               xend=xcrit,
               y=0,
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pnorm(xcrit, mu, sigma, lower.tail=F),
               yend=pnorm(xcrit, mu, sigma, lower.tail=F),
               linetype=2) +
  ylab('P(X<=x)') +
  theme(aspect.ratio=1)
```


```{r, echo=F}
n <- 30
p <- 0.5
x <- seq(0, n, 1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[1]
ga1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x>xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x<=xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))+
  theme(aspect.ratio=1)

ga2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit, 
               xend=xcrit, 
               y=0, 
               yend=pbinom(xcrit, n, p, lower.tail=F), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pbinom(xcrit, n, p, lower.tail=F),
               yend=pbinom(xcrit, n, p, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[2]
gb1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x<=xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x>xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

gb2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit, 
               xend=xcrit, 
               y=0, 
               yend=pbinom(xcrit, n, p, lower.tail=T), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit,
               y=pbinom(xcrit, n, p, lower.tail=T),
               yend=pbinom(xcrit, n, p, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=F)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[3]
gc1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x>=xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x<xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2))+
  theme(aspect.ratio=1)

gc2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit-1, 
               xend=xcrit-1, 
               y=0, 
               yend=pbinom(xcrit-1, n, p, lower.tail=F), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit-1,
               y=pbinom(xcrit-1, n, p, lower.tail=F),
               yend=pbinom(xcrit-1, n, p, lower.tail=F),
               linetype=2) +
  ylab('P(X>x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

fxd <- dbinom(x, n, p)
fxp <- pbinom(x, n, p, lower.tail=T)
d <- data.table(x, fxd, fxp)
xcrit <- xcritvec[4]
gd1b <- ggplot(d, aes(x, fxd)) +
  geom_segment(data=d[x<xcrit], aes(x=x, xend=x, y=0, yend=fxd), colour='blue') +
  geom_segment(data=d[x>=xcrit], aes(x=x, xend=x, y=0, yend=fxd)) +
  ylab('f(x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)

gd2b <- ggplot(d, aes(x, fxp)) +
  geom_point() +
  geom_segment(x=xcrit-1, 
               xend=xcrit-1, 
               y=0, 
               yend=pbinom(xcrit-1, n, p, lower.tail=T), 
               linetype=2) +
  geom_segment(x=min(x),
               xend=xcrit-1,
               y=pbinom(xcrit-1, n, p, lower.tail=T),
               yend=pbinom(xcrit-1, n, p, lower.tail=T),
               linetype=2) +
  ylab('P(X<=x)') +
  scale_x_continuous(breaks=seq(0,n,2)) +
  theme(aspect.ratio=1)
```

```{r, echo=F, fig.height=25}
plot_list <- list(ga1n, gd2n, gc2b, gb1n,
                  gd1b, gc1b, gb2n, ga2b)
plot_labels <- list('I', 'II', 'III', 'IV',
                    'V', 'VI', 'VII', 'VIII')
# plot_list <- list(ga1n, gb1n, gc1n, gd1n, 
#                   ga2n, gb2n, gc2n, gd2n,
#                   ga1b, gb1b, gc1b, gd1b, 
#                   ga2b, gb2b, gc2b, gd2b) 
ggarrange(plotlist=plot_list,
          labels=plot_labels,
          nrow=8,
          ncol=2)
```


## Problem 4

Consider the random variable $X{\sim}F(d1,d2)$ for $d1=5$,
$d2=100$.  For these values of $d1$ and $d2$,

$E[X] = \frac{d2}{d2 - 2}$

$Var[X] = \frac{2 d2^2 (d1 + d2 - 2)}{d1 (d2 - 2)^2 (d2 - 4)}$

**(a)** Compute the population mean and variance for $X$

