---
title: "Homework 8"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r, echo=F, message=F, warning=F}
library(data.table)
library(ggplot2)
library(ggpubr)
rm(list = ls())

gen_exp <- function(muA1, muA2, muB1, muB2, muC1, muC2, sig, n) {

  ## generate obervations
  y_A1 <- rnorm(n, muA1, sig)
  y_A2 <- rnorm(n, muA2, sig)
  y_B1 <- rnorm(n, muB1, sig)
  y_B2 <- rnorm(n, muB2, sig)
  y_C1 <- rnorm(n, muC1, sig) # new level but same mu
  y_C2 <- rnorm(n, muC2, sig) # new level but same mu

  ## store observations plus relevant indicators in data.table
  d <- data.table(y=c(y_A1, y_A2, y_B1, y_B2, y_C1, y_C2),
                  treatment=rep(c('A', 'B', 'C'), each=2*n),
                  dose=rep(c(1, 2), 3, each=n))

  d[, treatment := factor(treatment)]
  d[, dose := factor(dose)]

  ## main effect of treatment (A vs B)
  dd <- d[, .(mean(y), sd(y)/sqrt(length(unique(dose)))), .(treatment)]
  g1 <- ggplot(dd, aes(treatment, V1)) +
    geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
    theme_classic() +
    theme(aspect.ratio=1) +
    ylab('Observed mean effect') +
    ggtitle('Main effect')

  ## main effect of dose (1 vs 2)
  dd <- d[, .(mean(y), sd(y)/sqrt(length(unique(treatment)))), .(dose)]
  g2 <- ggplot(dd, aes(dose, V1)) +
    geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
    theme_classic() +
    theme(aspect.ratio=1) +
    ylab('Observed mean effect') +
    ggtitle('Main effect')

  ## interaction between treatment and dose
  dd <- d[, .(mean(y), sd(y)/sqrt(length(unique(treatment))*length(unique(dose)))), .(treatment, dose)]
  g3 <- ggplot(dd, aes(dose, V1, colour=treatment)) +
    geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
    geom_line(aes(as.integer(dose), V1, colour=treatment)) +
    theme_classic() +
    theme(aspect.ratio=1) +
    ylab('Observed mean effect') +
    ggtitle('Interaction')

  return(list(g1, g2, g3, unique(d)))
}
```

**1.** Considering the following data and using an
appropriate ANOVA, test the hypothesis that there are any
significant differences between mean levels of treatment.
Please populate the following variables.

```{r, echo=F}
set.seed(1)
n <- 4
res <- gen_exp(10, 30, 40, 20, 40, 20, 5, n)
g1 <- res[[1]]
g2 <- res[[2]]
g3 <- res[[3]]
d <- res[[4]]
d[, dose := NULL]
d[, subject := 1:.N]
d
```

```{r, eval=F}
# the result of ezANOVA()
prob_1_r_result <- ezANOVA(...)
```

**2.** Considering the following data and using an
appropriate ANOVA, test the hypothesis that there are any
significant differences between mean levels of treatment.
Please populate the following variables.

```{r, echo=F}
set.seed(1)
n <- 4
res <- gen_exp(10, 30, 40, 20, 40, 20, 5, n)
g1 <- res[[1]]
g2 <- res[[2]]
g3 <- res[[3]]
d <- res[[4]]
d[, dose := NULL]
d[, subject := rep(1:(2*n), 3)]
d
```

```{r, eval=F}
# the result of ezANOVA()
prob_2_r_result <- ezANOVA(...)
```

**3.** Considering the following data and using an
appropriate ANOVA, test the hypothesis that there are any
significant differences between mean levels of treatment,
mean levels of dose, or interactions between treatment and
dose. Please populate the following variables.

```{r, echo=F}
set.seed(1)
n <- 4
res <- gen_exp(10, 30, 40, 20, 40, 20, 5, n)
g1 <- res[[1]]
g2 <- res[[2]]
g3 <- res[[3]]
d <- res[[4]]
d[, subject := rep(1:(2*n), 3)]
d
```

```{r, eval=F}
# the result of ezANOVA()
prob_3_r_result <- ezANOVA(...)
```

**4.** Considering the following data and using an
appropriate ANOVA, test the hypothesis that there are any
significant differences between mean levels of treatment,
mean levels of dose, or interactions between treatment and
dose. Please populate the following variables.

```{r, echo=F}
set.seed(1)
n <- 4
res <- gen_exp(10, 30, 40, 20, 40, 20, 5, n)
g1 <- res[[1]]
g2 <- res[[2]]
g3 <- res[[3]]
d <- res[[4]]
d[, subject := 1:.N]
d
```

```{r, eval=F}
# the result of ezANOVA()
prob_4_r_result <- ezANOVA(...)
```

**5.** Consider the data located here:

https://crossley.github.io/cogs2020/data/maze/maze.csv

This data contains the results of an experiment in which
researchers examined the effect of maze identity and run
number (experience with each maze) on maze run time in a
large sample of rats.

* `rat` contains the id of the rat that gave the data in
each row.

* `y` contains the amount of time in seconds taken to
complete the maze.

* `maze` indicates the identity of the maze.

* `run` indicates the run number.

Using an appropriate ANOVA, assess any significant
differences between runs. If you have multiple observations
per rat, per run, then be sure to average over these values
such that you end up with only one (mean) observation per
rat per run.

```{r, eval=F}
# You do not have to perform the two-way ANOVA steps
# manually. Please use ezANOVA() and store the result of
# this function to a variable named `ans_3`
ans_5 <- ezANOVA(...)
```

**6.** Consider the following sampling distributions of two
$F$ test statistics. One of these sampling distributions
corresponds to a repeated measures design, while the other
corresponds to a factorial design. The observed value of the
F test statistic (indicated by the dashed lines) is 1.5 in
both cases.

```{r, echo=F, fig.width=10}
set.seed(7)

n_total <- 24
n <- 8
k <- 3

f_obs <- 1.5

# repeated
df_between <- k-1
df_error <- (k-1) * (n-1)
# pf(f_obs, df_between, df_error, lower.tail=F)

x <- seq(0, 4, 0.001)
fx <- df(x, df_between, df_error)
d <- data.table(x, fx)
g1 <- ggplot(d, aes(x, fx)) + 
  geom_line() +
  geom_vline(xintercept = f_obs, linetype=2) +
  scale_x_continuous(breaks=c(0, f_obs, 1, 2, 3, 4)) +
  geom_ribbon(data=d[x >= f_obs], aes(ymin=0, ymax=fx), alpha=0.25) +
  annotate("text", x = 1.6, y = 0.1, hjust = 0, label = "Area of shaded region = 0.26") 

# factorial
df_between <- k-1
df_error <- n_total - k
# pf(f_obs, df_between, df_error, lower.tail=F)

x <- seq(0, 4, 0.001)
fx <- df(x, df_between, df_error)
d <- data.table(x, fx)
g2 <- ggplot(d, aes(x, fx)) + 
  geom_line() +
  geom_vline(xintercept = f_obs, linetype=2) +
  scale_x_continuous(breaks=c(0, f_obs, 1, 2, 3, 4)) +
  geom_ribbon(data=d[x >= f_obs], aes(ymin=0, ymax=fx), alpha=0.25) +
  annotate("text", x = 1.6, y = 0.1, hjust = 0, label = "Area of shaded region = 0.25")

ggarrange(g1, g2, ncol=2)
```

* Which distribution corresponds to the repeated measures
design, and which to the factorial design?
  * `ans_6a <- "repeated on the left, factorial on the right"` 
  * `ans_6a <- "factorial on the left, repeated on the right"` 
  
* What is the p-value of the test corresponding the left
distribution? Please store your answer in a variable named
`ans_6b`.