---
title: "Lecture 12 - Comparing two treatments"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 2
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
editor_options:
  chunk_output_type: console
---

# Comparing two treatments

## Independent samples

Thus far we have learned how to test hypotheses that ask if
a sample of data were likely generated by a random variable
with a mean of some given value. For example, given sample
$x_1 \ldots x_n$ from a random variable $X$, do you believe
that $\mu_X > c$ where $c$ is some number. This situation
will occur in your research career numerous times, so
learning how to do this will easily prove to be a wise
investment. However, in many circumstances, we will want to
test hypotheses that ask if two samples were likely obtained
by different random variables or not. That is, we will want
to compare two treatments. The good news is that the
fundamental logic of NHST that we have already established
still works perfectly in this scenario.

As an example, lets return to the criterion learning data we
have discussed numerous times in lecture and homework.

```{r}

library(data.table)
library(ggplot2)
rm(list = ls())

d <- fread(
  'https://crossley.github.io/cogs2020/data/criterion_learning/crit_learn.csv')

d <- d[cnd %in% c('Delay', 'Short ITI', 'Long ITI')]

dd <- d[order(cnd, sub), mean(t2c), .(cnd, sub)]

ggplot(dd, aes(cnd, V1)) +
  geom_violin() +
  ylab('Mean Trials to Criterion')

```

Consider two random variables $X \sim \mathcal{N}(\mu_X,
\sigma_X)$ and $Y \sim \mathcal{N}(\mu_Y, \sigma_Y)$. Let
$X$ generate data for the *Delay* condition and $Y$ generate
data for the *Long ITI* condition. Furthermore, lets assume
that **$X$ and $Y$ are independent**. This means that given
a sample from $X$, you don't change your beliefs about $Y$
in any way, and vice versa.

Is it realistic to assume that $X$ and $Y$ are independent?
Since different subjects are in each condition (i.e., this
is a *between-subjects* design), we can assume independence.
We'd need to rethink our assumptions if the same subjects
were involved in sampling from both $X$ and $Y$.

This particular experiment was designed to answer a simple
question: **Does a feedback delay cause the mean trials to
criterion to increase** relative to a control condition
where feedback is given immediately and a delay is put in
the inter-trial interval.

<b>1\. Specify the null and alternative hypotheses ($H_0$ and
   $H_1$) in terms of a distribution and population
   parameter.</b>
$$
H_0: \mu_X = \mu_Y \\
H_1: \mu_X > \mu_Y
$$

<b>2\. Specify the type I error rate -- denoted by the symbol
   $\alpha$ -- you are willing to tolerate.</b>
   
$$\alpha = 0.05$$

<b>3\. Specify the sample statistic that you will use to
   estimate the population parameter in step 1 and state how
   it is distributed under the assumption that $H_0$ is
   true.</b>
   
Hmm. Here things seem to get a bit tricky. How do we
estimate $\mu_X = \mu_Y$? The first step is to rewrite our
hypotheses such that all parameters are on the same side of
the equation.

$$
H_0: \mu_X - \mu_Y = 0  \\
H_1: \mu_X - \mu_Y >  0
$$

The next step is define a third random variable $W = X - Y$. With this
definition, we know:

$$
W \sim \mathcal{N}(\mu_W, \sigma_W) \\
\mu_W = \mu_X - \mu_Y \\
\sigma^2_W = \sigma^2_X + \sigma^2_Y
$$

Now we can rewrite our hypotheses as:

$$
H_0: \mu_W = 0 \\
H_1: \mu_W > 0
$$

Finally, we can actually get to business as usual for step 3:

$$
\begin{align}
\widehat{\mu_W} &= \overline{W} \sim \mathcal{N}(\mu_\overline{W}, \sigma_\overline{W}) \\\\
\mu_\overline{W} &= \mu_\overline{X} - \mu_\overline{Y} \\
            &= \mu_X - \mu_Y \\\\
\sigma^2_\overline{W} &= \sigma^2_\overline{X} + \sigma^2_\overline{Y} \\
                 &= \frac{\sigma^2_X}{n} + \frac{\sigma^2_Y}{n}
\end{align}
$$

We know that $\sigma^2_W = \sigma^2_X + \sigma^2_Y$, so if
we somehow now $\sigma^2_X$ and $\sigma^2_Y$, then we can
just plug those in a proceed with Normal distributions. Much
more commonly, we will not know $\sigma^2_X$ and
$\sigma^2_Y$, so we will have to estimate them, and because
of this, $\hat{\mu_W}$ will have a $t$ distribution, not a
Normal distribution.

To figure out exactly how this t-distribution is configured,
we need to figure out (1) how to estimate $\sigma_\overline{W}$,
and (2) the degrees of freedom of the resulting $t$
distribution.

<b>3\. Specify the sample statistic that you will use to
estimate the population parameter in step 1 and state how it
is distributed under the assumption that $H_0$ is true.</b>

### Equal sample size and equal variance
$$
\begin{align}
s^2_\overline{W} &= s^2_\overline{X} + s^2_\overline{Y} \\
            &= \frac{s^2_X}{n_X} + \frac{s^2_Y}{n_Y} \\
            &= \frac{s^2_X}{n} + \frac{s^2_Y}{n} \\
            &= \frac{s^2_X + s^2_Y}{n} \\\\
s_\overline{W} &= \sqrt{\frac{s^2_X + s^2_Y}{n}}
\end{align}
$$

Here, it is common to define a variable called the **pooled
variance** denoted by $s_p$ and defined by:

$$
s^2_p = \frac{S^2_X + S^2_Y}{2}
$$

That is, in this case, the pooled variance is just the
average variance over each component. In any case, we can
define our test statistic and sampling distribution as
follows:

$$
\begin{align}
t_{obs} &= \frac{\overline{W} - \mu_\overline{W}}{s_\overline{W}} \\
        &= \frac{\overline{X} - \overline{Y}}{\sqrt{\frac{s^2_X + s^2_Y}{n}}} \\\\
t_{obs} &\sim t(df) \\\\
df &= (n_X-1) + (n_Y-1) \\
   &= 2n - 2
\end{align}
$$

Alternatively, you can write this using the pooled variance
as follows:

$$
\begin{align}
t_{obs} &= \frac{\overline{W} - \mu_\overline{W}}{s_\overline{W}} \\
        &= \frac{\overline{X} - \overline{Y}}{s_p \sqrt{\frac{2}{n}}} \\\\
t_{obs} &\sim t(df) \\\\
df &= (n_X-1) + (n_Y-1) \\
   &= 2n - 2
\end{align}
$$

### Unequal sample sizes and equal variances

With unequal sample sizes, the pooled variance is weighted
more towards the sample with the larger size:

$$
s^2_p = \frac{(n_X-1)s^2_X + (n_Y-1)s^2_Y}{n_X + n_Y - 2}
$$

The rest works out almost the same as above:

$$
\begin{align}
t_{obs} &= \frac{\overline{W} - \mu_\overline{W}}{s_\overline{W}} \\
        &= \frac{\overline{X} - \overline{Y}}{s_p \sqrt{\frac{1}{n_X} + \frac{1}{n_Y}}} \\\\
t_{obs} &\sim t(df) \\\\
df &= (n_X-1) + (n_Y-1) \\
   &= n_X + n_Y - 2
\end{align}
$$

### Unequal sample size and unequal variance

It turns out that the expression for our observed t-value and estimated variance
is quite simple, but the degrees of freedom for the test looks a bit more
complicated.

$$
\begin{align}
t_{obs} &= \frac{\overline{X} - \overline{Y}}{\sqrt{\frac{s^2_X}{n_X} + \frac{s^2_Y}{n_Y}}}\\\\
df &= \frac{\left( \frac{s^2_X}{n_X} + \frac{s^2_Y}{n_Y} \right)^2}
           {\frac{\left(\frac{s^2_X}{n_X} \right)^2}{n_X-1} + \frac{\left( \frac{s^2_Y}{n_Y} \right)^2}{n_Y-1}}
\end{align}
$$

This expression for $df$ is known as the Welch–Satterthwaite
equation.

<b>4\. Obtain a random sample and use it to compute the
sample statistic from step 3. Call this value
$\widehat{\theta}_{\text{obs}}$</b>

```{r}
x <- dd[cnd=='Delay', V1]
y <- dd[cnd=='Long ITI', V1]

nx <- length(x)
ny <- length(y)

xbar <- mean(x)
ybar <- mean(y)

varx <- var(x)
vary <- var(y)

## we have almost equal n, and lets assume equal variance
s <- sqrt((varx + vary)/nx)

tobs <- (xbar - ybar) / s
df <- nx + ny - 2
```

<b>5\. If $\widehat{\theta}_{\text{obs}}$ is very unlikely
to occur under the assumption that $H_0$ is true, then
reject $H_0$. Otherwise, do not reject $H_0$.</b>

```{r}
pval <- pt(tobs, df, lower.tail=F)

if(pval < 0.05) {
  print('reject the null')
  print(tobs)
  print(df)
  print(pval)
} else {
  print('fail to reject the null')
  print(tobs)
  print(df)
  print(pval)
}

## Check our work using built in t.test()
t.test(x=x,
       y=y,
       alternative='greater',
       mu=0,
       paired=FALSE,
       var.equal=TRUE,
       conf.leve=0.95)
```

Our results are very close to the results from `t.test()`
but not exact.

```{r}
## Don't assume equal sample sizes, but still assume equal variance
s <- sqrt( (((nx-1)*varx + (ny-1)*vary) / (nx + ny - 2)) * (1/nx + 1/ny) )
tobs <- (xbar - ybar) / s
df <- (varx/nx +vary/ny)^2 / ( (varx/nx)^2/(nx-1) + (vary/ny)^2/(ny-1))
pval <- pt(tobs, df, lower.tail=F)

if(pval < 0.05) {
  print('reject the null')
  print(tobs)
  print(df)
  print(pval)
} else {
  print('fail to reject the null')
  print(tobs)
  print(df)
  print(pval)
}

t.test(x=x,
       y=y,
       alternative='greater',
       mu=0,
       paired=FALSE,
       var.equal=TRUE,
       conf.leve=0.95)
```

Hooray! Now everything matches exactly!

## Paired samples

Suppose we ask the following question about our criterion
learning data:

- **Do people reach criterion faster on the second problem
than they do on the first problem?**

To address this question we proceed as we did in the
previous example. Let $X \sim \mathcal{N}(\mu_X, \sigma_X)$
and $Y \sim \mathcal{N}(\mu_Y, \sigma_Y)$. Let $X$ generate
data for *problem 1* condition and $Y$ generate data for the
*problem 2* condition.

In the previous example, $X$ and $Y$ were independent. This
was a reasonable assumption because our design is
between-subjects with respect to condition. Here, however,
we are comparing data from the same subjects on different
problems. That is, our design is within-subject with respect
to problem number. This means that we cannot assume
independence, and our procedure will look a bit different.

As we will see below, the trick will be to compute
difference scores for each subject, and then simply proceed
as we would with a one-sample t-test using these difference
scores as our sample.

<b>1\. Specify the null and alternative hypotheses ($H_0$
and $H_1$) in terms of a distribution and population
parameter.</b>
   
Let $D$ be the random variable that generates paired
difference scores between $X$ and $Y$.

$$
H_0: \mu_D = 0 \\
H_1: \mu_D > 0
$$

<b>2\. Specify the type I error rate -- denoted by the
symbol $\alpha$ -- you are willing to tolerate.</b>

$$\alpha = 0.05$$
   
<b>3\. Specify the sample statistic that you will use to
estimate the population parameter in step 1 and state how it
is distributed under the assumption that $H_0$ is true.</b>

$$
\begin{align}
\hat{\mu_D} = \overline{D} &= \frac{1}{n} \sum_{i=1}^n x_i - y_i \\
              \overline{D} &\sim \mathcal{N}(\mu_\overline{D}, \sigma_\overline{D}) \\
              \mu_\overline{D} &= \mu_D \\
              \sigma_\overline{D} &= \frac{\sigma_D}{\sqrt{n}}
\end{align}
$$

As usual, we will typically not know $\sigma_D$ and so we
will estimate it with the sample standard deviation, and the
resulting test statistic will be $t$.

$$
t_{obs} = \frac{\overline{D} - \mu_D}{s_D} \sim t(n - 1)
$$

where the degrees of freedom is $n-1$.

<b>4\. Obtain a random sample and use it to compute the
   sample statistic from step 3. Call this value
   $\widehat{\theta}_{\text{obs}}$</b>

```{r}
## Add a column to indicate the number of problems solved.
## This is important because we cant compute a diff score
## unless they solved at least two problems.
d[, nps := max(prob_num), .(cnd, sub)]

x <- d[nps > 1 & cnd %in% c('Delay', 'Long ITI') & prob_num==1, unique(t2c)]
y <- d[nps > 1 & cnd %in% c('Delay', 'Long ITI') & prob_num==2, unique(t2c)]

D <- x - y
n <- length(D)

Dbar <- mean(D)
Dbarsig <- sd(D) / sqrt(n)

tobs <- Dbar / Dbarsig
```

<b>5\. If $\widehat{\theta}_{\text{obs}}$ is very unlikely
to occur under the assumption that $H_0$ is true, then
reject $H_0$. Otherwise, do not reject $H_0$.</b>

```{r}
df <- n - 1
pval <- pt(tobs, df, lower.tail=F)

if(pval < 0.05) {
  print('reject the null')
  print(tobs)
  print(df)
  print(pval)
} else {
  print('fail to reject the null')
  print(tobs)
  print(df)
  print(pval)
}

# pass the difference scores and set paired to FALSE
t.test(x=D,
       alternative='greater',
       mu=0,
       paired=FALSE,
       var.equal=TRUE,
       conf.leve=0.95)

# pass both x and y and set paried to TRUE
t.test(x=x,
       y=y,
       alternative='greater',
       mu=0,
       paired=FALSE,
       var.equal=TRUE,
       conf.leve=0.95)
```
