---
title: "Midterm Prep"
author: "Author: Matthew J. Crossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 18pt
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
```

```{r setup, include=FALSE, message=F}
knitr::opts_chunk$set(echo = T, collapse=T, warning=F)
```

```{r echo=FALSE}
library(data.table)
library(ggplot2)
library(ggpubr)
library(latex2exp)

rm(list = ls())

source("util_funcs.R")
set.seed(0)
```

## PDF

### Practice 1

```{r echo=FALSE}

l <- generate_pdf_continuous()
g <- l[[1]]
d <- l[[2]]

g

## The values in the graph are given by the following:
x1 <- d[, x][1]
x2 <- d[, x][2]
x3 <- d[, x][3]
x4 <- d[, x][4]
x5 <- d[, xend][4]

y1 <- 0
y2 <- d[, sort(density)][1]
y3 <- d[, sort(density)][2]
y4 <- d[, sort(density)][3]
y5 <- d[, sort(density)][4]

```

- Is $X$ continuous or discrete?

- Compute $P(X < `r x3`)$

- Compute $P(X > `r x3`)$

- Compute $P(`r x1` \leq X < `r x3`)$

- Find $x$ such that $P(X < x) = 0.05$

- Find $x$ such that $P(X > x) = 0.05$

### Practice 2

```{r echo=FALSE}

l <- generate_pdf_discrete()
g <- l[[1]]
d <- l[[2]]

g

## The values in the graph are given by the following:
x1 <- d[, x][1]
x2 <- d[, x][2]
x3 <- d[, x][3]
x4 <- d[, x][4]

y1 <- d[, sort(mass)][1]
y2 <- d[, sort(mass)][2]
y3 <- d[, sort(mass)][3]
y4 <- d[, sort(mass)][4]

```

- Is $X$ continuous or discrete?

- Compute $P(X < `r x3`)$

- Compute $P(X > `r x3`)$

- Compute $P(`r x1` \leq X < `r x3`)$

- Compute the mean and variance of $X$

## CDF

### Practice 1

For the random variable $X \sim N(\mu_X=10, \sigma_X=2)$

- Compute $P(X < 12)$

- Compute $P(X > 8)$

- Compute $P(8 \leq X < 12)$

### Practice 2

```{r echo=FALSE}

l <- generate_cdf_discrete()

g <- l[[1]]
d <- l[[2]]

g

## The values in the graph are given by the following:
x1 <- d[, x][1]
x2 <- d[, x][2]
x3 <- d[, x][3]
x4 <- d[, x][4]

y1 <- d[, sort(probability)][1]
y2 <- d[, sort(probability)][2]
y3 <- d[, sort(probability)][3]
y4 <- d[, sort(probability)][4]

```

- Is $X$ continuous or discrete?

- Compute $P(X < `r x2`)$

- Compute $P(X \geq `r x2`)$

- Compute $P(`r x2` \leq X < `r x3`)$

## Quantile Function

### Practice 1

For the random variable $X \sim \mathcal{N}(\mu_x = 5, \sigma_x=3)$

- Find $x$ such that $P(X < x) = 0.5$

- Find $x$ such that $P(X < x) = 0.05$

- Find $x$ such that $P(X \geq x) = 0.05$

### Practice 2

For the random variable $X \sim \mathcal{Poisson}(\lambda=4)$

- Find $x$ such that $P(X \leq x) = 0.5$

- Find $x$ such that $P(X < x) = 0.05$

- Find $x$ such that $P(X > x) = 0.05$

## NHST

### Practice 1

Consider an experiment in which participants first learn novel categories, and
are subsequently asked to perform some other task. The question the researchers
are interested in is how initial learning affects the subsequent task, and it is
therefore critical that participants that didn't learn anything be excluded from
the final analysis. An exclusion criteria is set such that any participant that
fails to reach at least $60\%$ accuracy by the end of
training is excluded. Suppose that data is collected from 30
participants, and of these, 5 are excluded. Is this greater
than would be expected if the true probability -- denoted
$p$ -- of meeting the exclusion criteria was 0.1?

- Answer this question going through all 5 steps in long form.

- Is there a built in function you can use? If so, use it.

- Report a point estimate for $\hat{p}$.

- Report the critical value for $\alpha=0.05$.

- Compute power against the alternative $H_1$ that $p$ is $0.3$.

### Practice 2

```{r echo=FALSE}

n <- 10
mu_x <- 0.6
sigma_x <- 0.15
x <- rnorm(n, mu_x, sigma_x)
x[x > 1] <- 1

```

Consider an experiment in which participants study a list of
words and are subsequently asked to recall as many of the
studied words as possible. Suppose that data is collected
from `r n` participants and the percentage of words recalled
by each participant is given by

\begin{equation}
x = ( `r round(x, 2)` )
\end{equation}

Let $X$ be the random variable that each participants
performance is drawn from, and suppose we know $\sigma_x =
`r sigma_x`$.

Is $\mu_x$ significantly greater than $0.6$?

- Answer this question going through all 5 steps in long form.

- Is there a built in function you can use? If so, use it.

- Report point and interval estimates for $\hat{\mu_x}$.

- Report the critical value for $\alpha=0.05$.

- Compute power against the alternative $H_1$ that $\mu_x$ is $0.62$.


### Practice 3

```{r echo=FALSE}

n <- 15

mu_x <- 0
sigma_x <- 15
x <- rnorm(n, mu_x, sigma_x)

```

Consider an experiment in which participants attempt to toss
a ping pong ball as accurately as possible to a target.
Suppose that data is collected from `r n` participants and
that the mean reach error per participant is given by

\begin{equation}
x = ( `r round(x, 2)` )
\end{equation}

Let $X$ be the random variable that each participant's
performance is drawn from.

Is $\mu_x$ significantly different from $0$?

- Answer this question going through all 5 steps in long form.

- Is there a built in function you can use? If so, use it.

- Report point and interval estimates for $\hat{\mu_x}$.

- Report the critical value for $\alpha=0.05$.

- Compute power against the alternative $H_1$ that $\mu_x$
is $2$ assuming that the population standard deviation is
equal to the sample standard deviation.

## Data Wrangling

```{r echo=FALSE}

## Make data up -- Rats running a maze -- simple wide format
mu_x <- runif(1, 100, 100)
sigma_x <- runif(1, 100, 100)

maze_time_rat_1 <- rnorm(5, mu_x, sigma_x)
maze_time_rat_2 <- rnorm(5, mu_x, sigma_x)
maze_time_rat_3 <- rnorm(5, mu_x, sigma_x)
maze_time_rat_4 <- rnorm(5, mu_x, sigma_x)
maze_time_rat_5 <- rnorm(5, mu_x, sigma_x)

d_wide <- data.table(
  maze_time_rat_1,
  maze_time_rat_2,
  maze_time_rat_3,
  maze_time_rat_4,
  maze_time_rat_5
)

d_wide[, run := 1:.N]

## Make data up -- Rats running a maze -- simple long format
n_rats <- 5
n_runs_per_rat <- 5
n_rows <- n_rats*n_runs_per_rat
rat <- rep(1:n_rats, n_runs_per_rat)

run <- rep(1:n_runs_per_rat, each=n_rats)

mu_x <- runif(1, 100, 100)
sigma_x <- runif(1, 100, 100)
maze_time <- rnorm(n_rows,
                   mu_x,
                   sigma_x)

d_long <- data.table(run,
                     rat,
                     maze_time)

fwrite(d_wide, '../live_course_material/data/maze/d_wide.csv')
fwrite(d_long, '../live_course_material/data/maze/d_long.csv')

```

Consider an experiment in which a set of rats run through a
set of mazes some number of times each. The researchers
running this experiment are interested in how quickly rats
can run these mazes, and so they plan to record the time in
seconds each rat takes to complete each run of each maze.
Before investing time and resources into a full experiment,
two graduate students are tasked with running a scaled-down
version (i.e., pilot experiments) to make sure everything is
set up correctly.

* Graduate student 1 acquired the following data:

`https://crossley.github.io/cogs2020/data/maze/d_wide.csv`

* Graduate student 2 acquired the following data:

`https://crossley.github.io/cogs2020/data/maze/d_long.csv`

* After validating the pilot data, they ran a full
experiment and acquired the following data:

`https://crossley.github.io/cogs2020/data/maze/maze.csv`

### Practice 1

- Convert the following `data.table` acquired by graduate
student 1 (shown below) to long format.

```{r echo=FALSE}

d_wide

```

### Practice 2

- Convert the following `data.table` acquired by graduate
student 2 (shown below) to wide format.

```{r echo=FALSE}

d_long

```

### Practice 3

```{r echo=FALSE}

## Make data up -- Rats running a maze -- more complicated long format
n_rats <- round(runif(1, 20, 50))
n_mazes <- round(runif(1, 4, 10))
n_runs <- round(runif(1, 5, 20))
n_rows <- n_rats*n_mazes*n_runs

rat <- rep(1:n_rats, each=n_mazes*n_runs)
maze <- rep(1:n_mazes, times=n_rats, each=n_runs)
run <- rep(1:n_runs, times=n_rats*n_mazes)

time <- c()
for(i in 1:n_rats) {
  mu_x <- runif(1, 50, 200)
  sigma_x <- runif(1, 25, 50)
  for(j in 1:n_mazes) {
    mu_x <- mu_x + runif(1, -20, 20)
    sigma_x <- sigma_x + runif(1, -10, 10)
    time <- c(time, rnorm(n_runs, mu_x, sigma_x))
  }
}

d <- data.table(rat,
                maze,
                run,
                time)

```

```{r echo=FALSE}

fwrite(d, '../live_course_material/data/maze/maze.csv')

```

- Recreate the following visualisation of their results.
  Note that error bars represent standard error of the mean
  (SEM). Also note that to get SEM here, first compute the
  mean `time` per `maze` per `rat`. Then, compute the mean and
  standard error from these values.

```{r echo=FALSE}

dd <- d[, mean(time), .(rat, maze)]
ddd <- dd[, .(mean(V1), sd(V1)/sqrt(length(unique(rat)))), .(maze)]
ggplot(ddd, aes(maze, V1)) +
  geom_pointrange(aes(ymin=V1-V2, ymax=V1+V2)) +
  xlab('maze') +
  ylab('mean maze time')

```


## NHST concepts
```{r echo=F}

mux <- 20
sigx <- 5
x <- seq(-5*sigx + mux, 5*sigx + mux, 0.01)
fx <- dnorm(x, mux, sigx)
dtx <- data.table(x, fx)
gx <- ggplot(dtx, aes(x, fx)) + 
  geom_line() +
  xlab(TeX('$x$')) +
  ylab(TeX('Probability Density $f(X=x)$'))

n <- 15
muxbar <- 20
sigxbar <- 5 / sqrt(n)
xbar <- seq(-5*sigxbar + muxbar, 5*sigxbar + muxbar, 0.01)
fxbar <- dnorm(xbar, muxbar, sigxbar)
dtxbar <- data.table(xbar, fxbar)
gxbar <- ggplot(dtxbar, aes(xbar, fxbar)) + 
  geom_line() +
  xlab(TeX('$\\bar{x}$')) +
  ylab(TeX('Probability Density $f(\\bar{X}=\\bar{x})$'))

ggarrange(gx, gxbar, ncol=2)
```