---
title: "Midterm"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: false
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r echo=FALSE, message=F}
library(data.table)
library(ggplot2)
library(EnvStats)
rm(list = ls())
```

**0.** To initialise your file correctly, use the following
code chunk to start your script. Be sure to replace
`set.seed(0)` with `set.seed(YOUR_MQ_ID_NUMBER)` where
`YOUR_MQ_ID_NUMBER` is your MQ student ID as an integer. If
you do not already have `EnvStats` installed, then please
install it by executing `install.packages('EnvStats')` in
your console. Please do not include any `install.packages`
commands in the midterm .R file that you submit.

```{r, message=F}
library(data.table)
library(ggplot2)
library(EnvStats)

rm(list=ls())

my_name <- "ENTER YOUR NAME HERE AS A STRING"
my_student_id <-"ENTER YOUR STUDENT ID AS A STRING"

# replace 314159 with your student id number AS AN INTEGER 
id_as_integer <- 314159
```

--- 

**1.** Test the hypothesis that $\theta$ is non-zero, where
$\theta$ is a parameter of a random variable $X$. Ensure
that the type I error rate of your test is $\alpha=0.05$.

* Copy and run the following code chunk to create a plot of
the sampling distribution of $\widehat{\theta}$. The dashed
line in the resulting plot indicates the observed value of
$\widehat{\theta}$.

```{r, eval=F}
set.seed(id_as_integer)
x_max <- runif(1, min=1, max=100)
x_min <- -x_max
x_mode <- 0
x <- seq(2*x_min, 2*x_max, 0.001)
fx <- dtri(x, x_min, x_max, x_mode) + dunif(x, 2*x_min, 2*x_max)
fx <- fx / 2
d <- data.table(x, fx)
if(runif(1) > 0.5) {
  x_obs <- runif(1, min=x_min+x_min*0.1, max=x_mode-x_max*0.1)
} else {
  x_obs <- -runif(1, min=x_min+x_min*0.1, max=x_mode-x_max*0.1)
}
fx_obs <- dtri(x_obs, x_min, x_max, x_mode) + dunif(x_obs, 2*x_min, 2*x_max)
fx_obs <- fx_obs / 2
ggplot(d, aes(x, fx)) + 
  geom_line() +
  geom_segment(aes(x=2*x_min, xend=2*x_min, y=0, yend=min(fx))) + 
  geom_segment(aes(x=2*x_max, xend=2*x_max, y=0, yend=min(fx))) + 
  geom_vline(xintercept=x_obs, linetype=2) +
  scale_x_continuous(breaks=c(2*x_min, x_min, x_mode, x_max, 2*x_max, x_obs), 
                     minor_breaks=F) +
  scale_y_continuous(breaks=c(0, min(fx), fx_obs, max(fx)), 
                     minor_breaks=F, 
                     limits=c(0, max(fx))) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5))
```

* Create a variable named `p_val_1` and set its value to the
p-value of this test.

* Create a variable named `ci_lower_1` and set its value to
the lower confidence interval limit.

* Create a variable named `ci_upper_1` and set its value to
the upper confidence interval limit.

* Create a variable named `decision_1` and set its value to
`"reject"` if you reject the null hypothesis or to `"fail to
reject"` if you fail to reject the null.

--- 

**2.** Consider the random variable $X$ which is defined as a
process that yields the mean outcome across trials per
participant on some examination. Test the hypothesis that
$\mu_{X}$ is non-zero. Ensure that the type I error rate of
your test is $\alpha=0.05$.

* Copy and run the following code chunk to create a
`data.table` named `d` that contains the result of
performing an experiment in which `n_subs` participants were
examined on `n_trials`. The columns of `d` are as follows:

  * `subject` indicates the participant each row of data was
  observed from.
  
  * `trial` indicates the trial per participant each row of
  data was observed from.
  
  * `observation` indicates the value that was observed on
  each trial per participant.

```{r, eval=F}
set.seed(id_as_integer)
n_subs <- 10
n_trials <- 10
n_obs <- 1
subject <- rep(1:n_subs, each=n_trials)
trial <- rep(1:n_trials, n_subs)
mu <- runif(1, -1, 1)
sig <- runif(1, 0, 50)
observation <- rnorm(n_subs*n_trials*n_obs, mu, sig)
d <- data.table(subject, trial, observation)
```

* Create a variable named `p_val_2` and set its value to the
p-value of this test.

* Create a variable named `ci_lower_2` and set its value to
the lower confidence interval limit.

* Create a variable named `ci_upper_2` and set its value to
the upper confidence interval limit.

* Create a variable named `decision_2` and set its value to
`"reject"` if you reject the null hypothesis or to `"fail to
reject"` if you fail to reject the null.

--- 

**3.** Consider the random variable $X$ which is defined as a
process that yields the dichotomous outcome on each trial of
an examination named `exam_1`. Test the hypothesis that the
probability of success on each trial, $p$ is different than
$0.5$. Ensure that the type I error rate of your test is
$\alpha=0.05$.

* Copy and run the following code chunk to create a
`data.table` named `d` that contains the result of
performing an experiment in which `n_exams` examinations
each with `n_trials` trials was performed. The columns of
`d` are as follows:

  * `exam` indicates the name of the examination from which
  each row of data was observed from.
  
  * `trial` indicates the trial per participant each row of
  data was observed from.
  
  * `observation` indicates the value that was observed on
  each trial per participant. A success is coded by a one,
  and a failure is coded by a zero.

```{r, eval=F}
set.seed(id_as_integer)
exams <- c('exam_1', 'exam_2', 'exam_3')
n_exams <- length(exams)
n_trials <- 100
n_obs <- 1
exam <- rep(exams, each=n_trials)
trial <- rep(1:n_trials, n_exams)
observation <- rbinom(n_exams*n_trials*n_obs, 1, runif(1, 0.25, 0.75))
d <- data.table(exam, trial, observation)
```

* Create a variable named `p_val_3` and set its value to the
p-value of this test.

* Create a variable named `ci_lower_3` and set its value to
the lower confidence interval limit. Note that your result
may differ slightly from the result given by `binom.test`.
Here, I will be looking for the value obtained when using
the method taught in lecture, and covered in the midterm
prep and homework assignments.

* Create a variable named `ci_upper_3` and set its value to
the upper confidence interval limit. Note that your result
may differ slightly from the result given by `binom.test`.
Here, I will be looking for the value obtained when using
the method taught in lecture, and covered in the midterm
prep and homework assignments.

* Create a variable named `decision_3` and set its value to
`"reject"` if you reject the null hypothesis or to `"fail to
reject"` if you fail to reject the null.