---
title: "Midterm 2"
author: "Author: Matthew J. Crossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: false
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: false
    theme: cosmo
fontsize: 18pt
---

```{r echo=FALSE, message=F}
library(data.table)
library(ggplot2)
library(EnvStats)
rm(list = ls())
```

**0.** To initialise your file correctly, use the following
code chunk to start your script. Be sure to replace
`set.seed(0)` with `set.seed(YOUR_MQ_ID_NUMBER)` where
`YOUR_MQ_ID_NUMBER` is your MQ student ID as an integer. If
you do not already have `EnvStats` installed, then please
install it by executing `install.packages('EnvStats')` in
your console. Please do not include any `install.packages`
commands in the midterm .R file that you submit.

```{r, message=F}
library(data.table)
library(ggplot2)
library(EnvStats)

rm(list=ls())

my_name <- "ENTER YOUR NAME HERE AS A STRING"
my_student_id <-"ENTER YOUR STUDENT ID AS A STRING"

# replace 314159 with your student id number AS AN INTEGER 
id_as_integer <- 314159
```

--- 

**1.** Test the hypothesis that $\theta$ is non-zero, where
$\theta$ is a parameter of a random variable $X$. Ensure
that the type I error rate of your test is $\alpha=0.05$.

* Copy and run the following code chunk to create a plot of
the sampling distribution of $\widehat{\theta}$. The dashed
line in the resulting plot indicates the observed value of
$\widehat{\theta}$.

```{r, eval=F}
set.seed(id_as_integer)
offset <- runif(1, min=-100, max=100)
x <- round(seq(-3+offset, 3+offset, 1))
fx <- sample(1:length(x), size=length(x), replace=F)
fx <- fx / sum(fx)
d <- data.table(x, fx)
x_obs <- sample(x, 1)
ggplot(d, aes(x, fx)) + 
  geom_point() +
  geom_segment(aes(x=, xend=x, y=0, yend=fx)) +
  geom_vline(xintercept=x_obs, linetype=2) +
  scale_x_continuous(breaks=x,
                     minor_breaks=F) +
  scale_y_continuous(breaks=fx,
                     minor_breaks=F,
                     limits=c(0, max(fx)))
```

* Create a variable named `p_val_1` and set its value to the
p-value of this test.

* Create a variable named `ci_lower_1` and set its value to
the lower confidence interval limit.

* Create a variable named `ci_upper_1` and set its value to
the upper confidence interval limit.

* Create a variable named `decision_1` and set its value to
`"reject"` if you reject the null hypothesis or to `"fail to
reject"` if you fail to reject the null.

--- 

**2.** Consider an experiment in which, over the course of
several heats organised over multiple days, participants run
several 100 meter dashes as fast as they can. Consider the
random variable $X$ which is defined as a process that
yields the largest 100 meter dash time produced by a
participant across the entire experiment. Test the
hypothesis that $\mu_{X}$ is different than 40 seconds.
Ensure that the type I error rate of your test is
$\alpha=0.05$.

* Copy and run the following code chunk to create a
`data.table` named `d` that contains the result of the
experiment described above. The columns of `d` are as
follows:

  * `subject` indicates the participant each row of data was
  observed from.
  
  * `day` indicates the day of the races.
  
  * `heat` indicates the heat within each day.
  
  * `trial` indicates the order of the race within each day
  and heat.
  
  * `observation` indicates the value that was observed from
  each race per participant.

```{r, eval=F}
set.seed(id_as_integer)
n_subs <- 10
n_days <- 2
n_heats <- 3
n_trials <- 10
n_obs <- 1
subject <- rep(1:n_subs, each=n_days*n_heats*n_trials*n_obs)
day <- rep(1:n_days, n_subs, each=n_heats*n_trials*n_obs)
heat <- rep(1:n_heats, n_subs*n_days, each=n_trials*n_obs)
trial <- rep(1:n_trials, n_subs*n_days*n_heats, each=n_obs)
mu <- runif(1, 40, 60)
sig <- runif(1, 5, 10)
observation <- rnorm(n_subs*n_days*n_heats*n_trials*n_obs, mu, sig)
d <- data.table(subject, day, heat, trial, observation)
```

* Create a variable named `p_val_2` and set its value to the
p-value of this test.

* Create a variable named `ci_lower_2` and set its value to
the lower confidence interval limit.

* Create a variable named `ci_upper_2` and set its value to
the upper confidence interval limit.

* Create a variable named `decision_2` and set its value to
`"reject"` if you reject the null hypothesis or to `"fail to
reject"` if you fail to reject the null.

--- 

**3.** Consider the random variable $X$ which is defined as
a process that yields the number of false starts occurring
on day 2 of the experiment described in question 2 above.
Test the hypothesis that the probability of a false start on
day 2, $p$ is different than $0.5$. Ensure that the type I
error rate of your test is $\alpha=0.05$.

* Copy and run the following code chunk to create a
`data.table` named `d` that contains the result of the
experiment described above. The columns of `d` are as
follows:

  * `false_start` indicates whether or not there was a false
  start on a given trial.
  
  * All other columns are defined the same way that they
  were defined in problem 2.

```{r, eval=F}
set.seed(id_as_integer)
n_subs <- 10
n_days <- 2
n_heats <- 3
n_trials <- 10
n_obs <- 1
subject <- rep(1:n_subs, each=n_days*n_heats*n_trials*n_obs)
day <- rep(1:n_days, n_subs, each=n_heats*n_trials*n_obs)
heat <- rep(1:n_heats, n_subs*n_days, each=n_trials*n_obs)
trial <- rep(1:n_trials, n_subs*n_days*n_heats, each=n_obs)
mu <- runif(1, 40, 60)
sig <- runif(1, 5, 10)
observation <- rnorm(n_subs*n_days*n_heats*n_trials*n_obs, mu, sig)
d <- data.table(subject, day, heat, trial, observation)
d[, false_start := rbinom(n_subs*n_days*n_heats*n_trials*n_obs,  1,  0.5)]
```

* Create a variable named `p_val_3` and set its value to the
p-value of this test.

* Create a variable named `ci_lower_3` and set its value to
the lower confidence interval limit. Note that your result
may differ slightly from the result given by `binom.test`.
Here, I will be looking for the value obtained when using
the method taught in lecture, and covered in the midterm
prep and homework assignments.

* Create a variable named `ci_upper_3` and set its value to
the upper confidence interval limit. Note that your result
may differ slightly from the result given by `binom.test`.
Here, I will be looking for the value obtained when using
the method taught in lecture, and covered in the midterm
prep and homework assignments.

* Create a variable named `decision_3` and set its value to
`"reject"` if you reject the null hypothesis or to `"fail to
reject"` if you fail to reject the null.
