---
title: "Lecture 1 - Introduction to `R`"
author: "Author: Matthew J. Cossley"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: false
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    # code_folding: show
    number_sections: false
    theme: cosmo
fontsize: 14pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics
```{r}
# Use R as a calculator
2 + 2

# Store results in variables
x <- 2 + 2
x

# Perform mathematical operations on variables
x <- 2 + 2
y <- 3 + 7
x + y
```


## Functions
* Functions are objects that take *arguments* as inputs,
perform some operation on those inputs, and return the
results.

* `ls()` is a function that reports what objects (e.g.,
variables you ahve defined) are in the current environment
and therefore available for you to interact with.

```{r}
ls()
```

* `rm()` is a function that can remove objects from the
current environment.

* It takes several arguments. Type `?rm()` to see help
information on how to use it.

* We will commonly combine `ls()` and `rm()` to remove all
objects from the current environment. This is a good thing
to do at the beginning of every new script you write.

```{r}
# Remove all objects from the current R session.
rm(list = ls())
```

## Data types

### `numeric`
```{r}
x <- 2
class(x)
```

### `character`
```{r}
x <- 'a'
class(x)
```

### `logical` 
```{r}
x <- TRUE
class(x)
```

### `factor`

* A `factor` is a *categorical* data type. They are most
often used to code for different conditions in an
experiment.

```{r}
x <- factor(1)
class(x)
```

## Containers

### `vector`

* Create a vector with the function `c()`.

* The elements of a `vector` must be of the same type.

* Access element `i` of `vector` `x` with square brackets
(e.g., `x[i]`)

```{r}
# Create a vector with any three numbers you like.
x <- c(1, 2, 3.14159)
x

# Access the third element of x
x3 <- x[3]
x3
```


### `list`

* Create a list with the function `list()`

* The elements of a `list` can be of different types.

* Access element `i` of `list` `x` with *double* square
brackets (e.g., `x[[i]]`)

```{r}
# Create a three element list containing one numeric
# item, one `character` item, and one logical item.
x <- list(3.14159, 'pi', TRUE)
x

# Access the third element of x
x3 <- x[[3]]
x3
```


### `data.frame`

* Create a `data.frame` with the function `data.frame()`

* `data.frame` is pretty close what you might think of as an
excel spreadsheet.

* Access column `x` in `data.frame` `df` with the `$`
operator (e.g., `df$x`).

```{r}
# Create vectors to later store in a data frame
x <- c('I', 'I', 'I', 'II', 'II', 'II', 'III', 'III', 'III', 'IV', 'IV', 'IV')
y <- c('a', 'a', 'b', 'b', 'c', 'c', 'd', 'd', 'e', 'e', 'f', 'f')
z <- rnorm(12)

# Create the data frame
df <- data.frame(x, y, z)
df

# Access column x
df$x
```

### `data.table`

* `data.table` does everything `data.frame` does, and much more.

* Because `data.table` is so powerful and so fast, we won't
really bother much with the finer details of `data.frame`.

* To use `data.table` we must first load it into our
environment using the `library()` function. 

* If you get an error message complaining that `there is no
package called data.table` then install it by executing
`install.packages('data.table')`.

* Create a `data.table` with the `data.table()` function.

* General form: `DT[i, j, by]`

  * `i`: selects rows
  
  * `j`: specifies columns and operations to perform on those
  columns
  
  * `by`: specifies columns to group the operation performed
  in `j` by

```{r}
# Load the data.table library 
library(data.table)

# create a data table
dt <- data.table(x, y, z)
dt
```

#### Subset rows in `i`

* There are many ways to select rows.

  * Set `i` equal to a vector of integers
  
  * Set `i` equal to a logical expression
  
```{r}
# Select rows by passing integer indices
dt[c(2, 4)]

# Select rows by passing a logical expression
dt[x=='II']
```


#### Select columns with `j`

* First a quick note about specifying `i`.

* If you want to operate on every row of some set of
columns, leave the `i` argument blank.

* This will result in the first character inside the square
brackets being a comma, which looks a bit strange if you're
not used to it. Just remember to read it as *select every row*.

* To operate on specific columns, set `j` equal to a list
of the names of the columns you want.

* Inside the square brackets of a `data.table`, `list` can
be abbreviated with a `.`, which can also look strage at
first. Once you get used to it, you will appreciate the
brevity.
  
```{r}
# Select every row of columns y and z the cool way
dt[, .(y, z)]

# Select every row of columns y and z the long way
dt[, list(y, z)]
```


#### Operate on columns with `j`

* You can do a lot more than just return the columns
specified in `j`. In fact, you can perform any operation you
want on them.

```{r}
# return the mean and standard deviation of z
dt[, .(mean(z), sd(z))]
```


#### Combine `i` and `j`

* It is straightforward to combine `i` and `j`.

```{r}
# Select all rows for which x==II and return columns y and z
dt[x=='II', .(y, z)]
```


#### Group using `by`

* The real magic of `data.table` comes in the power of the
`by` argument.

* `by` allows you apply the operation that you specify with
`j` to separate groups of rows defined by the unique values
in the columns specified in `by`.

* Put another way, whatever column you pass to `by` will be
split up into groups with each unique value getting its own
group. Those groups will then be applied to the columns you
specify in `j` and the operation also specified in `j` will
be applied only to rows from the same group. Then, after all
is done, everything is put back into a single `data.table`.

```{r}
# Compute the sum of elements in column z separately for
# each value of x
dt[, sum(z), .(x)]
```

```{r}
# Compute the sum of elements in column z separately for
# each unique combination of values in x and y
dt[, sum(z), .(x, y)]
```


#### Adding and modifying columns

* Use the `:=` operator inside the `j` argument of a
`data.table` to add or modify a column.

```{r}
# If you pass a single number, data.table will fill the
# entire column with that value
dt[, a := 9]

# Otherwise, just pass a vector of the same length as the
# data.table
dt[, b := seq(2, 24, 2)]

# redefine or modify an existing column the same way
dt[, b := seq(4, 48, 4)]

# Remove a column using := and the NULL keyword
dt[, b := NULL]
```

#### Be careful when copying `data.table` objects

* `dt2 = dt` creates a *shallow copy*. This means that there
is really only one `data.table` object in memory, but it can
be referred to by both names.

* This means that changes to `dt2` will be reflected in `dt`
and vice-versa.

* You can show that this is the case by modifying dt2 and
observing that dt also changes (see code chunks below).

* To create a *deep copy* use `dt2 = data.table(dt)`.

```{r}
# shallow copy:
dt2 = dt # Not a deep copy
dt ## inspect dt before dt2 changes
dt2[, a := NULL]
dt ## inspect dt after dt changes

# deep copy:
dt2 = data.table(dt) # this is a deep copy
dt ## inspect dt before dt2 changes
dt2[, x := NULL]
dt ## inspect dt after dt changes
```

#### Converting from wide to long format

* Suppose you perform an experiment in which you collect
data from 3 participants, and each participant gives you 5
observations. 

* If this data were represented in *wide* format, then data
from each participant would be in separate columns.

* If this data were represented in *long* format, then data
from every participant would be in the same column. A
separate column would *indicate* what subject a given row
corresponded to.

* To convert from wide to long format use the `melt()`
function from `data.table`.

```{r}
x1 <- runif(5)
x2 <- runif(5)
x3 <- runif(5)
data_wide <- data.table(x1, x2, x3)
data_long <- melt(data_wide, measure.vars=c('x1', 'x2', 'x3'))

data_wide
data_long
```

